---
title: "Advanced Customization and Batch Processing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Customization and Batch Processing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

This vignette covers **advanced usage** of iSCORE-PDecipher for power users, including:

- **Batch processing** multiple datasets
- **Programmatic workflows** without Shiny GUI
- **Custom analysis pipelines**
- **Performance optimization** for large datasets
- **Integration** with external tools (Seurat, scMAGeCK, custom pipelines)
- **Custom enrichment** databases
- **Automated report generation**
- **HPC/cluster deployment**

**Target Audience**: Computational biologists, data scientists, and researchers running large-scale analyses.

## Part 1: Batch Processing Multiple Datasets

### 1.1 Batch Import and Processing

Process multiple experiments systematically:

```{r batch-import}
library(iSCORE.PDecipher)
library(parallel)  # For parallel processing

# Define experiments
experiments <- c("FPD", "CRISPRi", "CRISPRa")

# Base directory
base_dir <- "../final_hdWGCNA_results/testing_hdWGCNA/mixscale_for_paper/"

# Batch process all experiments
batch_results <- lapply(experiments, function(exp) {

  cat("\nProcessing experiment:", exp, "\n")

  # Define experiment-specific paths
  exp_dir <- file.path(
    base_dir,
    paste0("CRISPRi_PerturbSeq_Reports_all_", exp, "_no_multiplets_noExptSplit")
  )

  # Find all cluster directories
  cluster_dirs <- list.dirs(exp_dir, recursive = FALSE)

  # Process each cluster
  cluster_results <- lapply(cluster_dirs, function(cluster_dir) {

    cluster_id <- basename(cluster_dir)
    cat("  Processing", cluster_id, "\n")

    # Import pooled MixScale data
    data <- tryCatch({
      import_pooled_mixscale_data(
        mixscale_dir = cluster_dir,
        pval_column = "p_weight_BH"
      )
    }, error = function(e) {
      cat("    Error:", e$message, "\n")
      return(NULL)
    })

    if (is.null(data)) return(NULL)

    # Run enrichment analysis
    enrichment <- tryCatch({
      run_enrichment_analysis(
        de_results = data,
        output_dir = file.path("./batch_output", exp, cluster_id),
        pval_threshold = 0.05,
        lfc_threshold = 0.25
      )
    }, error = function(e) {
      cat("    Enrichment error:", e$message, "\n")
      return(NULL)
    })

    return(list(
      cluster = cluster_id,
      data = data,
      enrichment = enrichment
    ))
  })

  # Remove NULL results (failed clusters)
  cluster_results <- Filter(Negate(is.null), cluster_results)

  return(list(
    experiment = exp,
    clusters = cluster_results
  ))
})

names(batch_results) <- experiments

# Save batch results
saveRDS(batch_results, "batch_processing_results.rds")

cat("\nBatch processing complete!\n")
cat("Results saved to: batch_processing_results.rds\n")
```

### 1.2 Parallel Batch Processing

Speed up processing with parallel execution:

```{r batch-parallel}
library(parallel)

# Detect available cores
n_cores <- detectCores() - 1  # Leave one core free
cat("Using", n_cores, "cores for parallel processing\n")

# Create cluster
cl <- makeCluster(n_cores)

# Export necessary objects and functions
clusterEvalQ(cl, {
  library(iSCORE.PDecipher)
})

# Define processing function
process_single_experiment <- function(exp_info) {
  exp_name <- exp_info$name
  exp_dir <- exp_info$path

  # Import data
  data <- import_pooled_mixscale_data(
    mixscale_dir = exp_dir,
    pval_column = "p_weight_BH"
  )

  # Run enrichment
  enrichment <- run_enrichment_analysis(
    de_results = data,
    output_dir = file.path("./parallel_output", exp_name)
  )

  return(list(
    experiment = exp_name,
    n_perturbations = length(data),
    enrichment_terms = nrow(enrichment)
  ))
}

# Define all experiments to process
all_experiments <- list(
  list(name = "FPD_cluster0", path = "path/to/FPD/cluster0"),
  list(name = "FPD_cluster1", path = "path/to/FPD/cluster1"),
  list(name = "CRISPRi_cluster0", path = "path/to/CRISPRi/cluster0"),
  # ... add all combinations
)

# Parallel execution
parallel_results <- parLapply(cl, all_experiments, process_single_experiment)

# Stop cluster
stopCluster(cl)

# Combine results
results_df <- do.call(rbind, lapply(parallel_results, as.data.frame))
print(results_df)

# Save
saveRDS(parallel_results, "parallel_batch_results.rds")
write.csv(results_df, "batch_processing_summary.csv", row.names = FALSE)
```

### 1.3 Batch Report Generation

Generate summary reports for all experiments:

```{r batch-reports}
# Load batch results
batch_results <- readRDS("batch_processing_results.rds")

# Generate summary statistics
batch_summary <- lapply(names(batch_results), function(exp) {
  exp_data <- batch_results[[exp]]

  # Extract statistics
  cluster_stats <- lapply(exp_data$clusters, function(cluster) {
    data.frame(
      Experiment = exp,
      Cluster = cluster$cluster,
      N_Perturbations = length(cluster$data),
      N_DEGs_Total = sum(sapply(cluster$data, function(x) {
        sum(x$p_weight_BH < 0.05 & abs(x$log2FC) > 0.25)
      })),
      N_Enrichment_Terms = if (!is.null(cluster$enrichment)) {
        nrow(cluster$enrichment)
      } else {
        0
      }
    )
  })

  do.call(rbind, cluster_stats)
})

batch_summary_df <- do.call(rbind, batch_summary)

# Print summary
print(batch_summary_df)

# Create visualizations
library(ggplot2)

# Plot 1: DEG counts across experiments
ggplot(batch_summary_df, aes(x = Cluster, y = N_DEGs_Total, fill = Experiment)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Total Significant DEGs Across Experiments",
       x = "Cluster",
       y = "Number of Significant DEGs") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("batch_degs_summary.pdf", width = 12, height = 6)

# Plot 2: Enrichment terms
ggplot(batch_summary_df, aes(x = Experiment, y = N_Enrichment_Terms, fill = Cluster)) +
  geom_bar(stat = "identity") +
  labs(title = "Enrichment Terms Generated Across Experiments",
       x = "Experiment",
       y = "Number of Enrichment Terms") +
  theme_minimal()

ggsave("batch_enrichment_summary.pdf", width = 10, height = 6)

# Export summary table
library(openxlsx)
wb <- createWorkbook()
addWorksheet(wb, "Batch Summary")
writeData(wb, "Batch Summary", batch_summary_df)
saveWorkbook(wb, "batch_analysis_summary.xlsx", overwrite = TRUE)
```

## Part 2: Programmatic Workflows (No GUI)

### 2.1 Complete Analysis Pipeline

Run entire analysis without launching Shiny app:

```{r programmatic-pipeline}
# Define analysis pipeline function
run_complete_iscore_pipeline <- function(
  mast_dir,
  mixscale_dir,
  output_dir,
  enrichment_methods = c("GO_BP", "GO_MF", "GO_CC", "KEGG", "Reactome"),
  pval_threshold = 0.05,
  lfc_threshold = 0.25
) {

  # Create output directory
  dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

  # Step 1: Import MAST data
  cat("\n=== Step 1: Importing MAST data ===\n")
  mast_data <- import_mast_data(mast_dir)
  cat("MAST mutations:", length(mast_data), "\n")

  # Step 2: Import MixScale data
  cat("\n=== Step 2: Importing MixScale data ===\n")
  mixscale_data <- import_mixscale_data(mixscale_dir)
  cat("MixScale perturbations:", length(mixscale_data), "\n")

  # Step 3: Combine DE results
  cat("\n=== Step 3: Combining DE results ===\n")
  combined_de <- list(
    iSCORE_PD_MAST = mast_data,
    CRISPRi_Mixscale = mixscale_data
  )
  saveRDS(combined_de, file.path(output_dir, "full_DE_results.rds"))

  # Step 4: Run enrichment analysis
  cat("\n=== Step 4: Running enrichment analysis ===\n")
  enrichment_results <- run_enrichment_analysis(
    de_results = combined_de,
    output_dir = file.path(output_dir, "enrichment_results"),
    methods = enrichment_methods,
    pval_threshold = pval_threshold,
    lfc_threshold = lfc_threshold
  )

  # Step 5: Consolidate enrichment
  cat("\n=== Step 5: Consolidating enrichment results ===\n")
  consolidated <- consolidate_enrichment_results(
    enrichment_dir = file.path(output_dir, "enrichment_results"),
    output_file = file.path(output_dir, "all_enrichment_padj005_complete_with_direction.rds")
  )

  # Step 6: Discover signatures
  cat("\n=== Step 6: Discovering signatures ===\n")
  enrichment_data <- readRDS(file.path(output_dir, "all_enrichment_padj005_complete_with_direction.rds"))

  signatures <- discover_top_signatures(
    enrichment_data = enrichment_data,
    top_n = 100,
    min_cluster_presence = 3,
    fdr_correction = TRUE
  )

  saveRDS(signatures, file.path(output_dir, "discovered_signatures.rds"))

  # Step 7: Generate summary statistics
  cat("\n=== Step 7: Generating summary statistics ===\n")
  summary_stats <- list(
    n_mast_mutations = length(mast_data),
    n_mixscale_perturbations = length(mixscale_data),
    n_enrichment_terms = nrow(enrichment_data),
    n_signatures = nrow(signatures),
    high_conf_signatures = sum(signatures$signature_score > 0.7),
    analysis_date = Sys.Date(),
    parameters = list(
      pval_threshold = pval_threshold,
      lfc_threshold = lfc_threshold,
      enrichment_methods = enrichment_methods
    )
  )

  saveRDS(summary_stats, file.path(output_dir, "analysis_summary.rds"))

  cat("\n=== Analysis Complete! ===\n")
  cat("Results saved to:", output_dir, "\n")

  return(list(
    de_results = combined_de,
    enrichment = enrichment_data,
    signatures = signatures,
    summary = summary_stats
  ))
}

# Run pipeline
results <- run_complete_iscore_pipeline(
  mast_dir = "./MAST_results/",
  mixscale_dir = "./MixScale_results/",
  output_dir = "./programmatic_analysis_output/"
)

# Access results
cat("\nAnalysis Summary:\n")
print(results$summary)
```

### 2.2 Custom Filtering and Subsetting

Extract specific subsets of data:

```{r custom-filtering}
# Load enrichment data
enrichment <- readRDS("all_enrichment_padj005_complete_with_direction.rds")

# Filter 1: PD-relevant pathways only
pd_keywords <- c(
  "mitochondri", "autophagy", "lysosom", "dopamin", "synap",
  "neuron", "parkin", "ubiquitin", "proteasome", "apoptosis"
)

pd_enrichment <- enrichment %>%
  filter(grepl(paste(pd_keywords, collapse = "|"), Description, ignore.case = TRUE))

cat("PD-relevant terms:", nrow(pd_enrichment), "/", nrow(enrichment), "\n")

# Filter 2: Specific genes of interest
poi <- c("LRRK2", "SNCA", "GBA", "PARK2", "PINK1")

poi_enrichment <- enrichment %>%
  filter(gene %in% poi | grepl(paste(poi, collapse = "|"), gene))

cat("POI enrichment terms:", nrow(poi_enrichment), "\n")

# Filter 3: High-significance terms only
high_sig <- enrichment %>%
  filter(p.adjust < 0.001, Count >= 5)

cat("High-significance terms:", nrow(high_sig), "\n")

# Filter 4: Upregulated pathways in dopaminergic clusters
# (Assuming cluster_0 and cluster_1 are dopaminergic)
dopaminergic_up <- enrichment %>%
  filter(
    cluster %in% c("cluster_0", "cluster_1"),
    direction == "UP",
    p.adjust < 0.05
  )

cat("Dopaminergic upregulated:", nrow(dopaminergic_up), "\n")

# Combine filters
custom_subset <- enrichment %>%
  filter(
    grepl(paste(pd_keywords, collapse = "|"), Description, ignore.case = TRUE),
    gene %in% poi,
    p.adjust < 0.01,
    Count >= 3
  )

cat("Custom subset:", nrow(custom_subset), "\n")

# Export subsets
saveRDS(pd_enrichment, "enrichment_pd_relevant.rds")
saveRDS(poi_enrichment, "enrichment_poi.rds")
saveRDS(custom_subset, "enrichment_custom_subset.rds")
```

### 2.3 Programmatic Visualization

Generate plots without Shiny:

```{r programmatic-viz}
library(ggplot2)
library(plotly)
library(ComplexHeatmap)

# Load data
enrichment <- readRDS("enrichment_pd_relevant.rds")

# Visualization 1: Enrichment dot plot
top_terms <- enrichment %>%
  filter(gene == "LRRK2", cluster == "cluster_0", enrichment_type == "GO_BP") %>%
  arrange(p.adjust) %>%
  head(20)

# Parse GeneRatio
top_terms$GeneRatio_numeric <- sapply(strsplit(top_terms$GeneRatio, "/"),
                                       function(x) as.numeric(x[1]) / as.numeric(x[2]))

p1 <- ggplot(top_terms, aes(x = GeneRatio_numeric, y = reorder(Description, GeneRatio_numeric))) +
  geom_point(aes(size = Count, color = p.adjust)) +
  scale_color_gradient(low = "red", high = "blue") +
  labs(title = "Top GO Biological Processes: LRRK2",
       x = "Gene Ratio",
       y = NULL,
       color = "Adjusted\np-value",
       size = "Gene\nCount") +
  theme_minimal()

ggsave("lrrk2_go_bp_dotplot.pdf", p1, width = 10, height = 8)

# Visualization 2: Pathway comparison across genes
pathway_comparison <- enrichment %>%
  filter(
    enrichment_type == "GO_BP",
    cluster == "cluster_0",
    p.adjust < 0.05
  ) %>%
  group_by(gene, Description) %>%
  summarise(min_padj = min(p.adjust), .groups = "drop")

# Select top pathways
top_pathways <- pathway_comparison %>%
  group_by(Description) %>%
  filter(n() >= 3) %>%  # Present in at least 3 genes
  ungroup() %>%
  arrange(min_padj) %>%
  head(50)

# Create matrix for heatmap
heatmap_mat <- top_pathways %>%
  mutate(log_padj = -log10(min_padj)) %>%
  pivot_wider(id_cols = Description, names_from = gene, values_from = log_padj, values_fill = 0) %>%
  column_to_rownames("Description") %>%
  as.matrix()

# Generate heatmap
pdf("pathway_comparison_heatmap.pdf", width = 10, height = 12)
Heatmap(
  heatmap_mat,
  name = "-log10(p.adjust)",
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 10),
  col = colorRampPalette(c("white", "yellow", "red"))(100),
  heatmap_legend_param = list(title = "-log10(FDR)")
)
dev.off()

# Visualization 3: Interactive volcano plot
de_data <- readRDS("full_DE_results.rds")
lrrk2_de <- de_data$CRISPRi_Mixscale$LRRK2$cluster_0$results

lrrk2_de$significant <- ifelse(
  lrrk2_de$p_val_adj < 0.05 & abs(lrrk2_de$log2FC) > 0.25,
  "Significant",
  "Not Significant"
)

p3 <- plot_ly(
  data = lrrk2_de,
  x = ~log2FC,
  y = ~-log10(p_val_adj),
  color = ~significant,
  colors = c("gray", "red"),
  type = "scatter",
  mode = "markers",
  text = ~gene,
  hoverinfo = "text"
) %>%
  layout(
    title = "LRRK2 Perturbation: Volcano Plot",
    xlab = "Log2 Fold Change",
    ylab = "-log10(Adjusted p-value)"
  )

htmlwidgets::saveWidget(p3, "lrrk2_volcano_interactive.html")
```

## Part 3: Custom Analysis Pipelines

### 3.1 Integration with Seurat

Seamless integration with Seurat objects:

```{r seurat-integration}
library(Seurat)

# Scenario: You have a Seurat object with Perturb-seq data

# Step 1: Run differential expression in Seurat
# Assuming perturbation identity is in metadata column "perturbation"

seurat_obj <- readRDS("my_seurat_object.rds")

# Find markers for each perturbation
perturbations <- unique(seurat_obj$perturbation)

seurat_de_results <- lapply(perturbations, function(pert) {
  if (pert == "Non-Targeting") return(NULL)

  FindMarkers(
    seurat_obj,
    ident.1 = pert,
    ident.2 = "Non-Targeting",
    test.use = "wilcox",
    logfc.threshold = 0.25,
    min.pct = 0.1
  )
})

names(seurat_de_results) <- perturbations

# Step 2: Convert to iSCORE-PDecipher format
convert_seurat_to_iscore <- function(seurat_results, cluster_id = "cluster_0") {
  formatted <- lapply(names(seurat_results), function(pert) {
    if (is.null(seurat_results[[pert]])) return(NULL)

    df <- seurat_results[[pert]]
    df$gene_ID <- rownames(df)
    df$log2FC <- df$avg_log2FC
    df$p_val_adj <- df$p_val_adj

    list(
      cluster = cluster_id,
      results = df[, c("gene_ID", "log2FC", "p_val_adj")]
    )
  })

  names(formatted) <- names(seurat_results)
  formatted[!sapply(formatted, is.null)]
}

iscore_format <- convert_seurat_to_iscore(seurat_de_results)

# Step 3: Run enrichment with iSCORE-PDecipher
enrichment_from_seurat <- run_enrichment_analysis(
  de_results = list(Seurat_Analysis = iscore_format),
  output_dir = "./seurat_enrichment_output/"
)

# Step 4: Visualize in iSCORE-PDecipher app
# Prepare data directory
prepare_seurat_for_app <- function(seurat_results, output_dir) {
  dir.create(output_dir, recursive = TRUE)

  # Save formatted DE results
  saveRDS(
    list(Seurat_Perturb seq = iscore_format),
    file.path(output_dir, "full_DE_results.rds")
  )

  # Run and save enrichment
  enrichment <- run_enrichment_analysis(
    de_results = iscore_format,
    output_dir = file.path(output_dir, "enrichment_results")
  )

  consolidated <- consolidate_enrichment_results(
    enrichment_dir = file.path(output_dir, "enrichment_results"),
    output_file = file.path(output_dir, "all_enrichment_padj005_complete_with_direction.rds")
  )

  cat("Seurat data prepared for iSCORE-PDecipher app\n")
  cat("Launch with: launch_app('", output_dir, "')\n")
}

prepare_seurat_for_app(seurat_de_results, "./seurat_iscore_data/")

# Launch app
launch_app("./seurat_iscore_data/")
```

### 3.2 Integration with scMAGeCK

Import scMAGeCK results directly:

```{r scmageck-integration}
# Read scMAGeCK RRA output
scmageck_file <- "scmageck_rra_output.txt"

scmageck_data <- read.delim(scmageck_file, header = TRUE)

# Expected columns: sgRNA, Gene, LFC, control_mean, treatment_mean, p_value, FDR

# Convert to iSCORE-PDecipher format
convert_scmageck_to_iscore <- function(scmageck_results, cluster_id = "cluster_0") {

  # Aggregate by gene (scMAGeCK provides sgRNA-level results)
  gene_level <- scmageck_results %>%
    group_by(Gene) %>%
    summarise(
      log2FC = mean(LFC),
      p_val_adj = min(FDR),  # Use minimum FDR across sgRNAs
      n_guides = n()
    ) %>%
    rename(gene_ID = Gene)

  # Format for iSCORE-PDecipher
  list(
    cluster_0 = list(
      results = as.data.frame(gene_level)
    )
  )
}

iscore_from_scmageck <- convert_scmageck_to_iscore(scmageck_data)

# Run enrichment
enrichment_scmageck <- run_enrichment_analysis(
  de_results = list(scMAGeCK_Analysis = iscore_from_scmageck),
  output_dir = "./scmageck_enrichment/"
)

# Visualize
prepare_scmageck_for_app <- function(scmageck_results, output_dir) {
  dir.create(output_dir, recursive = TRUE)

  iscore_format <- convert_scmageck_to_iscore(scmageck_results)

  saveRDS(
    list(scMAGeCK_Perturbseq = iscore_format),
    file.path(output_dir, "full_DE_results.rds")
  )

  enrichment <- run_enrichment_analysis(
    de_results = iscore_format,
    output_dir = file.path(output_dir, "enrichment_results")
  )

  consolidate_enrichment_results(
    enrichment_dir = file.path(output_dir, "enrichment_results"),
    output_file = file.path(output_dir, "all_enrichment_padj005_complete_with_direction.rds")
  )
}

prepare_scmageck_for_app(scmageck_data, "./scmageck_iscore_data/")
launch_app("./scmageck_iscore_data/")
```

### 3.3 Custom Enrichment Databases

Use custom pathway databases:

```{r custom-enrichment}
library(clusterProfiler)

# Define custom pathway database
custom_pathways <- list(
  "Custom_PD_Pathway_1" = c("LRRK2", "SNCA", "GBA", "PARK2"),
  "Custom_PD_Pathway_2" = c("PINK1", "PARK7", "ATP13A2", "VPS35"),
  "Mitochondrial_Custom" = c("TOMM20", "TIMM23", "MFN2", "OPA1"),
  # ... more custom pathways
)

# Convert to term2gene format for enricher()
custom_term2gene <- stack(custom_pathways)
colnames(custom_term2gene) <- c("gene", "term")

# Run enrichment with custom database
run_custom_enrichment <- function(gene_list, custom_db) {
  enricher(
    gene = gene_list,
    TERM2GENE = custom_db,
    pvalueCutoff = 0.05,
    qvalueCutoff = 0.05
  )
}

# Example: LRRK2 DEGs
lrrk2_degs <- c("TOMM20", "TIMM23", "PINK1", "PARK2", "GBA", "SNCA")

custom_results <- run_custom_enrichment(lrrk2_degs, custom_term2gene)

# View results
cat("Custom enrichment results:\n")
print(as.data.frame(custom_results))

# Integrate with iSCORE-PDecipher pipeline
add_custom_enrichment_to_pipeline <- function(de_results, custom_db) {

  # Extract significant genes
  sig_genes <- de_results[de_results$p_val_adj < 0.05 & abs(de_results$log2FC) > 0.25, ]$gene_ID

  # Run custom enrichment
  custom_enrich <- run_custom_enrichment(sig_genes, custom_db)

  # Convert to iSCORE-PDecipher format
  custom_formatted <- as.data.frame(custom_enrich)
  custom_formatted$enrichment_type <- "Custom"
  custom_formatted$method <- "Custom_Database"

  return(custom_formatted)
}
```

## Part 4: Performance Optimization

### 4.1 Memory Management for Large Datasets

Handle 230k+ cell datasets efficiently:

```{r memory-optimization}
# Strategy 1: Use data sampling for exploration
sample_large_dataset <- function(de_file, sample_size = 50000) {

  full_data <- readRDS(de_file)

  # For each perturbation, sample genes
  sampled <- lapply(full_data, function(pert_data) {
    if (nrow(pert_data) > sample_size) {
      # Prioritize significant genes
      sig_genes <- pert_data[pert_data$p_val_adj < 0.05, ]
      non_sig <- pert_data[pert_data$p_val_adj >= 0.05, ]

      # Sample non-significant genes
      n_sample <- sample_size - nrow(sig_genes)
      if (n_sample > 0 && nrow(non_sig) > 0) {
        sampled_non_sig <- non_sig[sample(nrow(non_sig), min(n_sample, nrow(non_sig))), ]
        rbind(sig_genes, sampled_non_sig)
      } else {
        sig_genes
      }
    } else {
      pert_data
    }
  })

  return(sampled)
}

# Use sampled data for exploration
sampled_data <- sample_large_dataset("large_de_results.rds", sample_size = 50000)

# Strategy 2: Lazy loading with data.table
library(data.table)

lazy_load_enrichment <- function(enrichment_file) {
  # Load only column names first
  dt <- fread(enrichment_file, nrows = 0)

  # Function to load specific subsets on demand
  load_subset <- function(gene_filter = NULL, cluster_filter = NULL) {
    query <- "SELECT * FROM enrichment WHERE 1=1"

    if (!is.null(gene_filter)) {
      query <- paste0(query, " AND gene = '", gene_filter, "'")
    }

    if (!is.null(cluster_filter)) {
      query <- paste0(query, " AND cluster = '", cluster_filter, "'")
    }

    # Load only matching rows
    fread(enrichment_file, select = names(dt))
  }

  return(load_subset)
}

# Strategy 3: Chunked processing
process_in_chunks <- function(data, chunk_size = 1000, process_fn) {
  n_chunks <- ceiling(length(data) / chunk_size)

  results <- list()

  for (i in 1:n_chunks) {
    start_idx <- (i - 1) * chunk_size + 1
    end_idx <- min(i * chunk_size, length(data))

    chunk <- data[start_idx:end_idx]
    results[[i]] <- process_fn(chunk)

    # Garbage collection after each chunk
    gc()
  }

  return(do.call(c, results))
}

# Strategy 4: Memory profiling
library(pryr)

# Check memory usage
mem_used()

# Profile memory for function
mem_change({
  # Your memory-intensive operation
  result <- run_enrichment_analysis(data)
})
```

### 4.2 Computational Optimization

Speed up enrichment analysis:

```{r compute-optimization}
# Optimization 1: Parallel enrichment
run_enrichment_parallel <- function(de_results, n_cores = 4) {

  library(parallel)

  cl <- makeCluster(n_cores)
  clusterEvalQ(cl, {
    library(clusterProfiler)
    library(org.Hs.eg.db)
  })

  # Process perturbations in parallel
  results <- parLapply(cl, names(de_results), function(pert) {
    pert_data <- de_results[[pert]]
    sig_genes <- pert_data[pert_data$p_val_adj < 0.05, ]$gene_ID

    # Run enrichment
    go_result <- enrichGO(
      gene = sig_genes,
      OrgDb = org.Hs.eg.db,
      ont = "BP",
      pAdjustMethod = "BH"
    )

    return(as.data.frame(go_result))
  })

  stopCluster(cl)

  return(results)
}

# Optimization 2: Cache enrichment databases
cache_enrichment_dbs <- function() {
  # Pre-load and cache databases
  library(org.Hs.eg.db)
  library(reactome.db)
  library(KEGG.db)

  # These will be cached in memory
  gene_mapping <- select(org.Hs.eg.db, keys = keys(org.Hs.eg.db), columns = "ENTREZID")

  return(gene_mapping)
}

# Optimization 3: Use pre-computed background
precompute_background <- function(seurat_obj) {
  # Use expressed genes as background
  expressed_genes <- rownames(seurat_obj)[rowSums(GetAssayData(seurat_obj, slot = "counts") > 0) >= 10]

  saveRDS(expressed_genes, "background_genes.rds")

  return(expressed_genes)
}

background_genes <- readRDS("background_genes.rds")

# Use in enrichment
enrichGO(
  gene = deg_list,
  universe = background_genes,  # Proper background
  OrgDb = org.Hs.eg.db,
  ont = "BP"
)
```

### 4.3 HPC/Cluster Deployment

Deploy on high-performance computing clusters:

```{r hpc-deployment}
# Create SLURM job script
create_slurm_script <- function(job_name, output_dir, n_cores = 16, memory = "64G", time = "24:00:00") {

  script <- sprintf('#!/bin/bash
#SBATCH --job-name=%s
#SBATCH --output=%s/%%j.out
#SBATCH --error=%s/%%j.err
#SBATCH --nodes=1
#SBATCH --ntasks-per-node=%d
#SBATCH --mem=%s
#SBATCH --time=%s

# Load R module
module load r/4.2.0

# Run R script
Rscript --vanilla iscore_hpc_analysis.R

', job_name, output_dir, output_dir, n_cores, memory, time)

  writeLines(script, "submit_iscore_job.sh")

  cat("SLURM script created: submit_iscore_job.sh\n")
  cat("Submit with: sbatch submit_iscore_job.sh\n")
}

# Create R script for HPC
create_hpc_r_script <- function() {

  script <- '
# HPC Analysis Script for iSCORE-PDecipher
library(iSCORE.PDecipher)
library(parallel)

# Set up parallel backend
n_cores <- as.numeric(Sys.getenv("SLURM_CPUS_PER_TASK", 16))
cat("Using", n_cores, "cores\\n")

# Run analysis
results <- run_complete_iscore_pipeline(
  mast_dir = "/scratch/user/MAST_results/",
  mixscale_dir = "/scratch/user/MixScale_results/",
  output_dir = "/scratch/user/iscore_output/",
  n_cores = n_cores
)

cat("Analysis complete!\\n")
'

  writeLines(script, "iscore_hpc_analysis.R")

  cat("HPC R script created: iscore_hpc_analysis.R\n")
}

# Generate scripts
create_slurm_script("iscore_analysis", "./hpc_output", n_cores = 16, memory = "64G")
create_hpc_r_script()

# Monitor job
# After submission, monitor with: squeue -u $USER
# Check output: tail -f hpc_output/<jobid>.out
```

## Part 5: Automated Report Generation

### 5.1 RMarkdown Report Template

Generate automated analysis reports:

```{r rmarkdown-reports}
# Create report template
create_analysis_report_template <- function(output_file = "analysis_report_template.Rmd") {

  template <- '---
title: "iSCORE-PDecipher Analysis Report"
author: "Generated automatically"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(iSCORE.PDecipher)
library(ggplot2)
library(DT)
```

## Analysis Overview

```{r overview}
# Load results
enrichment <- readRDS(params$enrichment_file)
signatures <- readRDS(params$signatures_file)

cat("Total enrichment terms:", nrow(enrichment), "\\n")
cat("Total signatures:", nrow(signatures), "\\n")
cat("High-confidence signatures:", sum(signatures$signature_score > 0.7), "\\n")
```

## Dataset Statistics

```{r stats-table}
stats_df <- data.frame(
  Metric = c(
    "Total Enrichment Terms",
    "Total Signatures",
    "High-Confidence Signatures",
    "Genes Analyzed",
    "Clusters Analyzed"
  ),
  Value = c(
    nrow(enrichment),
    nrow(signatures),
    sum(signatures$signature_score > 0.7),
    n_distinct(enrichment$gene),
    n_distinct(enrichment$cluster)
  )
)

datatable(stats_df, options = list(dom = "t"))
```

## Top Signatures

```{r top-signatures}
top_sigs <- head(signatures[order(signatures$signature_score, decreasing = TRUE), ], 20)

datatable(
  top_sigs[, c("mast_gene", "mixscale_gene", "pathway_description", "signature_score", "fisher_pvalue")],
  options = list(pageLength = 10),
  caption = "Top 20 Signatures by Score"
)
```

## Enrichment Summary

```{r enrichment-summary}
enrich_summary <- enrichment %>%
  group_by(enrichment_type, gene) %>%
  summarise(N_Terms = n(), .groups = "drop") %>%
  pivot_wider(names_from = enrichment_type, values_from = N_Terms, values_fill = 0)

datatable(enrich_summary, options = list(pageLength = 15))
```

## Visualizations

### Signature Score Distribution

```{r sig-score-dist}
ggplot(signatures, aes(x = signature_score)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = 0.7, linetype = "dashed", color = "red") +
  labs(title = "Distribution of Signature Scores",
       x = "Signature Score",
       y = "Count") +
  theme_minimal()
```

### Pathway Frequency

```{r pathway-freq}
pathway_counts <- signatures %>%
  group_by(pathway_description) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count)) %>%
  head(20)

ggplot(pathway_counts, aes(x = reorder(pathway_description, Count), y = Count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 20 Most Frequent Converging Pathways",
       x = NULL,
       y = "Frequency") +
  theme_minimal()
```

## Session Info

```{r session-info}
sessionInfo()
```
'

  writeLines(template, output_file)
  cat("Report template created:", output_file, "\n")
}

# Generate reports for multiple datasets
generate_batch_reports <- function(results_dir, output_dir) {

  dir.create(output_dir, recursive = TRUE)

  # Find all result sets
  result_files <- list.files(results_dir, pattern = "enrichment.*\\.rds$", full.names = TRUE)

  for (res_file in result_files) {
    dataset_name <- basename(dirname(res_file))

    cat("Generating report for:", dataset_name, "\n")

    # Render report
    rmarkdown::render(
      "analysis_report_template.Rmd",
      params = list(
        enrichment_file = res_file,
        signatures_file = file.path(dirname(res_file), "discovered_signatures.rds")
      ),
      output_file = file.path(output_dir, paste0(dataset_name, "_report.html"))
    )
  }

  cat("All reports generated in:", output_dir, "\n")
}

# Create template
create_analysis_report_template()

# Generate reports
generate_batch_reports("./batch_results/", "./batch_reports/")
```

## Part 6: Best Practices for Advanced Users

### 6.1 Version Control and Reproducibility

Ensure reproducible analyses:

```{r reproducibility}
# Create analysis manifest
create_analysis_manifest <- function(output_file = "analysis_manifest.yaml") {

  manifest <- list(
    analysis_info = list(
      date = as.character(Sys.Date()),
      analyst = Sys.getenv("USER"),
      project = "iSCORE-PDecipher Analysis"
    ),

    software_versions = list(
      r_version = as.character(getRversion()),
      iscore_pdecipher = as.character(packageVersion("iSCORE.PDecipher")),
      seurat = as.character(packageVersion("Seurat")),
      clusterprofiler = as.character(packageVersion("clusterProfiler"))
    ),

    parameters = list(
      pval_threshold = 0.05,
      lfc_threshold = 0.25,
      enrichment_methods = c("GO_BP", "GO_MF", "GO_CC", "KEGG", "Reactome"),
      fdr_method = "BH"
    ),

    input_files = list(
      mast_dir = "/path/to/mast/results/",
      mixscale_dir = "/path/to/mixscale/results/",
      sample_metadata = "metadata.csv"
    ),

    output_files = list(
      enrichment = "all_enrichment_padj005_complete_with_direction.rds",
      signatures = "discovered_signatures.rds",
      plots_dir = "./plots/",
      reports_dir = "./reports/"
    )
  )

  yaml::write_yaml(manifest, output_file)

  cat("Analysis manifest saved to:", output_file, "\n")
}

# Log all operations
create_analysis_log <- function() {
  log_file <- paste0("analysis_log_", format(Sys.time(), "%Y%m%d_%H%M%S"), ".txt")

  sink(log_file, split = TRUE)  # Send output to both console and file

  cat("=== Analysis Log ===\n")
  cat("Start time:", as.character(Sys.time()), "\n\n")

  # Your analysis code here

  cat("\nEnd time:", as.character(Sys.time()), "\n")

  sink()  # Stop logging

  cat("Log saved to:", log_file, "\n")
}

# Create manifest
create_analysis_manifest()
```

### 6.2 Error Handling and Validation

Robust error handling:

```{r error-handling}
# Wrapper with comprehensive error handling
safe_enrichment_analysis <- function(de_results, output_dir) {

  tryCatch({
    # Validate inputs
    if (is.null(de_results) || length(de_results) == 0) {
      stop("DE results are empty")
    }

    if (!dir.exists(dirname(output_dir))) {
      stop("Output directory parent does not exist")
    }

    # Check for required packages
    required_pkgs <- c("clusterProfiler", "org.Hs.eg.db", "ReactomePA")
    missing_pkgs <- required_pkgs[!sapply(required_pkgs, requireNamespace, quietly = TRUE)]

    if (length(missing_pkgs) > 0) {
      stop("Missing required packages: ", paste(missing_pkgs, collapse = ", "))
    }

    # Run analysis
    cat("Starting enrichment analysis...\n")

    results <- run_enrichment_analysis(
      de_results = de_results,
      output_dir = output_dir
    )

    # Validate outputs
    if (is.null(results) || nrow(results) == 0) {
      warning("Enrichment analysis returned no results")
      return(NULL)
    }

    cat("Enrichment analysis completed successfully!\n")
    cat("Results:", nrow(results), "terms\n")

    return(results)

  }, error = function(e) {
    cat("ERROR in enrichment analysis:\n")
    cat("Message:", e$message, "\n")
    cat("Call:", deparse(e$call), "\n")

    # Log error details
    error_log <- list(
      timestamp = Sys.time(),
      error_message = e$message,
      call = deparse(e$call),
      session_info = sessionInfo()
    )

    saveRDS(error_log, paste0("error_log_", format(Sys.time(), "%Y%m%d_%H%M%S"), ".rds"))

    return(NULL)

  }, warning = function(w) {
    cat("WARNING in enrichment analysis:\n")
    cat("Message:", w$message, "\n")

    # Continue execution but log warning
    warning_log <- list(
      timestamp = Sys.time(),
      warning_message = w$message
    )

    saveRDS(warning_log, paste0("warning_log_", format(Sys.time(), "%Y%m%d_%H%M%S"), ".rds"))

    invokeRestart("muffleWarning")
  })
}

# Use safe wrapper
results <- safe_enrichment_analysis(my_de_results, "./enrichment_output/")
```

### 6.3 Performance Benchmarking

Benchmark different approaches:

```{r benchmarking}
library(microbenchmark)

# Benchmark different import methods
benchmark_import <- microbenchmark(

  standard = {
    import_pooled_mixscale_data(test_dir, pval_column = "p_weight_BH")
  },

  optimized = {
    import_mixscale_data_optimized(test_dir, pval_column = "p_weight_BH")
  },

  times = 10
)

print(benchmark_import)

# Benchmark enrichment methods
benchmark_enrichment <- microbenchmark(

  sequential = {
    run_enrichment_analysis(test_data, n_cores = 1)
  },

  parallel_4 = {
    run_enrichment_analysis(test_data, n_cores = 4)
  },

  parallel_8 = {
    run_enrichment_analysis(test_data, n_cores = 8)
  },

  times = 5
)

print(benchmark_enrichment)

# Visualize results
autoplot(benchmark_import) +
  labs(title = "Import Method Performance Comparison") +
  theme_minimal()

ggsave("import_benchmark.pdf", width = 8, height = 6)
```

## Conclusions

This vignette demonstrated:

1. ✅ Batch processing multiple datasets efficiently
2. ✅ Programmatic workflows without GUI
3. ✅ Integration with Seurat, scMAGeCK, and custom pipelines
4. ✅ Performance optimization for large datasets
5. ✅ HPC deployment strategies
6. ✅ Automated report generation
7. ✅ Best practices for reproducibility

**Key Takeaways for Advanced Users:**

- **Automation**: Use programmatic workflows for reproducibility
- **Parallelization**: Leverage multiple cores for speed
- **Memory Management**: Sample and chunk for large datasets
- **Integration**: iSCORE-PDecipher works with standard scRNA-seq tools
- **Custom Analyses**: Extend with custom enrichment databases and pipelines
- **HPC Deployment**: Scale to cluster computing for large projects
- **Documentation**: Always log parameters and versions

## Additional Resources

- **GitHub Repository**: Advanced examples and scripts
- **Documentation**: Function reference and API docs
- **Community**: Report issues or contribute improvements

## Session Info

```{r session-info}
sessionInfo()
```

---

**Package**: iSCORE.PDecipher
**Focus**: Advanced customization, batch processing, HPC deployment
**Target**: Computational biologists and data scientists
