---
title: "Getting Started with iSCORE-PDecipher"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with iSCORE-PDecipher}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

**iSCORE-PDecipher** is an R package for integrated analysis of Parkinson's disease research data, combining:

- **iSCORE-PD Genetic Mutations**: 13 PD-associated mutations across 14 cell clusters
- **CRISPRi Gene Knockdowns**: 10 PD genes across multiple cell clusters (Perturb-seq)
- **Comprehensive Enrichment**: ~663,000+ significant enrichment terms
- **Interactive Visualization**: Shiny application for data exploration

This vignette provides a complete guide to installing, configuring, and using iSCORE-PDecipher for your first analysis.

## Installation

### System Requirements

- **R version**: ≥ 4.0.0
- **Platform**: Windows, macOS, or Linux
- **Memory**: At least 8GB RAM recommended (16GB+ for large datasets)

### Platform-Specific Prerequisites

**Windows:**
- Install [RTools](https://cran.r-project.org/bin/windows/Rtools/)

**macOS:**
- Install Xcode command line tools: `xcode-select --install`

**Linux:**
- Install build-essential: `sudo apt-get install build-essential`

### Install Required Bioconductor Packages

```{r install-bioconductor}
# Install BiocManager if needed
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

# Install required Bioconductor packages
BiocManager::install(c(
  "clusterProfiler",
  "ReactomePA",
  "DOSE",
  "org.Hs.eg.db",
  "pathview",
  "SingleCellExperiment",
  "dittoSeq",
  "enrichplot",
  "ComplexHeatmap"
))
```

### Install CRAN Packages

```{r install-cran}
# Install CRAN packages
install.packages(c(
  "heatmaply",
  "colourpicker",
  "shinyWidgets",
  "shinycssloaders",
  "shinyjs",
  "ggridges"
))
```

### Install iSCORE-PDecipher

```{r install-package}
# Install from GitHub (recommended)
if (!require("remotes", quietly = TRUE))
  install.packages("remotes")

remotes::install_github("jessedunnack/iSCORE-PDecipher")
```

**Installation time**: 5-15 minutes depending on internet speed and system.

## Quick Start

### Load the Package

```{r load-package}
library(iSCORE.PDecipher)
```

### First Launch: Interactive Setup

On your first launch, iSCORE-PDecipher will guide you through a one-time configuration:

```{r first-launch}
# Launch the interactive app
launch_app()

# First-time setup prompts:
# 1. Welcome message
# 2. Prompt for parent data directory location
# 3. Dataset selection menu
# 4. App launches with selected dataset
```

**What happens during first launch:**

1. **Configuration Check**: The package checks if you've launched before
2. **Parent Directory Setup**: You'll be asked where your data is stored
3. **Configuration Saved**: Settings are saved for future sessions (platform-appropriate location)
4. **Dataset Selection**: Choose from pre-configured datasets or provide custom path

### Dataset Selection

After initial setup, you'll see a dataset selection menu:

```
=== Select Dataset ===
Available datasets:
[1] iSCORE-PD + CRISPRi Combined Dataset
    /path/to/iSCORE-PD_plus_CRISPRi/
[2] iSCORE-PD Only Dataset
    /path/to/iSCORE-PD/
[3] CRISPRi Only Dataset
    /path/to/CRISPRi/
[C] Choose custom directory
[Q] Quit

Enter your choice:
```

**Type the number** of your desired dataset and press Enter.

## Understanding the Application Interface

### Main Dashboard Layout

Once launched, the iSCORE-PDecipher app has these main tabs:

1. **Overview**: Landing page with UMAP visualization and dataset statistics
2. **DE Results**: Interactive differential expression results with volcano plots
3. **Enrichment**: Browse and filter enrichment terms
4. **Heatmaps**: Generate interactive and static heatmaps
5. **Signature Nomination**: Discover convergent pathways between mutations and perturbations
6. **Signature Trends**: Data-driven trend analysis
7. **Export**: Export data and figures

### Global Filter Panel (Left Sidebar)

The collapsible left sidebar contains global filters that apply across all modules:

- **Gene/Mutation**: Select specific genes or mutations to analyze
- **Cluster**: Filter by cell cluster (cluster_0 through cluster_N)
- **Enrichment Type**: Choose GO_BP, GO_CC, GO_MF, KEGG, Reactome, WikiPathways, STRING, or GSEA
- **Direction**: Filter by gene regulation direction (ALL, UP, DOWN, BOTH)
- **Analysis Method**: Select MAST (mutations), MixScale (perturbations), or both
- **P-value Threshold**: Set significance threshold (default: 0.05)

**Tip**: These filters update all visualizations in real-time!

## Basic Analysis Workflow

### Step 1: Explore the Overview

The **Overview** tab provides:

- **UMAP Plot**: Interactive visualization of cell clusters (click clusters to filter)
- **Dataset Statistics**:
  - Total genes analyzed
  - Number of cell clusters
  - Number of enrichment terms
- **Cluster Markers**: Top marker genes for each cluster

**Try this:**
- Click on a cluster in the UMAP plot
- Observe how the statistics update
- Check the marker genes table for that cluster

### Step 2: Examine Differential Expression Results

Navigate to the **DE Results** tab:

**Features:**
- **Dual Volcano Plots**: MAST (left) and MixScale (right) side-by-side
- **Interactive Points**: Hover to see gene names and statistics
- **Gene Highlighting**: Select genes to highlight across both plots
- **DE Table**: Comprehensive results table below plots

**Workflow:**
```
1. Use global filters to select:
   - Gene/Mutation: "LRRK2"
   - Cluster: "cluster_0"

2. Examine volcano plots:
   - Red points: Upregulated (log2FC > threshold, p < threshold)
   - Blue points: Downregulated (log2FC < -threshold, p < threshold)
   - Gray points: Not significant

3. Hover over points to see:
   - Gene name
   - Log2 fold change
   - Adjusted p-value

4. Click gene names in table to highlight in plots

5. Download results using "Download DE Results" button
```

### Step 3: Browse Enrichment Terms

Navigate to the **Enrichment** tab:

**Features:**
- **Enrichment Table**: All significant terms (p.adjust < threshold)
- **Gene List Display**: Click term to see associated genes
- **Comparison View**: Compare terms across conditions

**Workflow:**
```
1. Filter enrichment terms:
   - Gene: "LRRK2"
   - Cluster: "cluster_0"
   - Enrichment Type: "GO_BP"
   - Direction: "UP"

2. Sort table by p.adjust (lowest first)

3. Click on a pathway (e.g., "mitochondrial transport")

4. View genes in that pathway:
   - Gene symbols
   - Expression changes
   - Links to NCBI/Ensembl

5. Copy gene list to clipboard or download
```

### Step 4: Generate Heatmaps

Navigate to the **Heatmaps** tab:

**Heatmap Types:**
- **Interactive** (heatmaply): Hover, zoom, pan
- **Static** (ComplexHeatmap): Publication-quality PDF export

**Workflow:**
```
1. Select data for heatmap:
   - Data Type: "P-values" (or Fold Enrichment, Z-scores, NES)
   - Genes/Mutations: Select multiple (e.g., LRRK2, SNCA, GBA)
   - Clusters: Select multiple (e.g., cluster_0, cluster_1, cluster_2)
   - Terms: Select pathways of interest or use "Top N" option

2. Configure visualization:
   - Cluster Rows: Yes
   - Cluster Columns: Yes
   - Distance: "euclidean"
   - Linkage: "complete"

3. Customize colors:
   - Low value: Blue
   - Mid value: White
   - High value: Red
   - Scaling: "Row" (recommended for p-values)

4. Generate heatmap

5. Export:
   - Interactive: Download HTML
   - Static: Download PDF (300+ DPI)
   - Data: Download matrix as CSV
```

**Interpreting the Heatmap:**
- **Rows**: Enrichment terms (pathways)
- **Columns**: Genes/Mutations across clusters
- **Colors**:
  - Red: High significance (low p-value or high enrichment)
  - Blue: Low significance
  - White: Neutral/medium
- **Dendrograms**: Show hierarchical relationships

### Step 5: Discover Signatures

Navigate to the **Signature Nomination** tab:

**What are signatures?**
Signatures are **convergent pathways** - biological processes affected by BOTH a genetic mutation (MAST) AND a gene knockdown perturbation (CRISPRi). These represent robust, cross-method validated findings.

**Workflow:**
```
1. Select gene pair:
   - MAST Gene (Mutation): "LRRK2_G2019S"
   - MixScale Gene (Perturbation): "LRRK2"
   - Click "Discover Signatures"

2. Review signature results table:
   - Pathway: Name of enriched term
   - Overlap genes: Genes affected by both
   - Fisher p-value: Statistical significance of overlap
   - Effect correlation: Correlation of effect sizes
   - Signature score: Composite score (higher = stronger signature)

3. Explore PD Biology Focus tab:
   - See biological categories (mitochondrial, autophagy, etc.)
   - View PD-relevant pathways
   - Read interpretation text

4. Check Pan-Cluster Signatures:
   - Pathways consistent across multiple clusters
   - Adjust "Minimum Clusters" slider

5. Export signatures for further analysis
```

**Interpreting Signatures:**
- **High signature score (>0.7)**: Strong convergent pathway
- **Low Fisher p-value (<0.01)**: Significant gene overlap
- **High correlation (>0.5)**: Similar effect directions

## Working with Your Own Data

### Scenario 1: You Have Pre-Computed Enrichment Results

If you already have enrichment results from another pipeline:

```{r custom-data-precomputed}
# Place your data in the expected directory structure:
# my_dataset/
# ├── full_DE_results.rds
# └── all_enrichment_padj005_complete_with_direction.rds

# Launch with your dataset
launch_app(data_dir = "/path/to/my_dataset/")
```

**Required file formats:**

**full_DE_results.rds** structure:
```r
list(
  iSCORE_PD_MAST = list(
    mutation1 = list(
      cluster_0 = list(results = data.frame(gene, log2FC, p_val_adj, ...)),
      cluster_1 = list(results = data.frame(...))
    )
  ),
  CRISPRi_Mixscale = list(
    gene1 = list(
      cluster_0 = list(results = data.frame(...))
    )
  )
)
```

**all_enrichment_padj005_complete_with_direction.rds** structure:
```r
data.frame(
  gene = character,           # Gene or mutation name
  cluster = character,        # Cluster ID
  enrichment_type = character, # GO_BP, KEGG, etc.
  Description = character,    # Pathway description
  GeneRatio = character,      # e.g., "15/200"
  BgRatio = character,        # e.g., "100/20000"
  pvalue = numeric,
  p.adjust = numeric,
  qvalue = numeric,
  geneID = character,         # Gene IDs in pathway
  Count = integer,            # Number of genes
  direction = character,      # "UP", "DOWN", or "ALL"
  method = character          # "MAST" or "MixScale"
)
```

### Scenario 2: You Have Raw MAST/MixScale Results

If you have raw differential expression results:

```{r custom-data-raw}
# Run the complete analysis pipeline
results <- run_complete_pipeline(
  mast_directory = "./my_MAST_results/",
  mixscale_directory = "./my_MixScale_results/",
  output_directory = "./my_iscore_output/",
  launch_app = TRUE  # Automatically launch when done
)

# The pipeline will:
# 1. Import MAST and MixScale results
# 2. Run enrichment analysis (GO, KEGG, Reactome, etc.)
# 3. Consolidate results
# 4. Validate dataset
# 5. Launch the Shiny app
```

### Scenario 3: You Have Perturb-seq Data Only

If you have Perturb-seq data without mutations (NEW in v0.5.0+):

```{r perturbseq-only}
# Import pooled MixScale data with FDR correction
mixscale_data <- import_pooled_mixscale_data(
  mixscale_dir = "./perturbseq_results/",
  pval_column = "p_weight_BH"  # Use Benjamini-Hochberg FDR correction
)

# Import corresponding enrichment
enrichment <- import_enrichment_with_correction(
  dataset = "my_perturbseq",
  pval_correction = "BH"
)

# Launch app
launch_app(data_dir = "./prepared_perturbseq_data/")
```

**P-value correction options:**
- `"p_weight"`: Uncorrected p-values
- `"p_weight_BH"`: Benjamini-Hochberg FDR (recommended)
- `"p_weight_bonferroni"`: Bonferroni correction (very conservative)

## Exporting Results

### Export from Shiny App

Navigate to the **Export** tab:

**Data Export Options:**
1. **Filtered Data**: Current filter settings
   - DE Results (CSV, TSV, RDS)
   - Enrichment Terms (CSV, Excel)
   - Signature Results (RDS, CSV)
   - Gene Lists (TXT)

2. **Figures**: Publication-quality exports
   - Volcano Plots (PDF, PNG, SVG)
   - Heatmaps (PDF, HTML)
   - UMAP Plots (PDF, PNG)
   - Resolution: 300-600 DPI

3. **Complete Dataset**: Full analysis (RDS)

### Programmatic Export

Export without using the Shiny interface:

```{r programmatic-export}
# Load data
enrichment <- readRDS("all_enrichment_padj005_complete_with_direction.rds")

# Filter for specific analysis
library(dplyr)
lrrk2_results <- enrichment %>%
  filter(gene == "LRRK2", cluster == "cluster_0", p.adjust < 0.05)

# Export to CSV
write.csv(lrrk2_results, "LRRK2_cluster0_enrichment.csv", row.names = FALSE)

# Export to Excel with multiple sheets
library(openxlsx)
wb <- createWorkbook()

addWorksheet(wb, "GO_BP")
writeData(wb, "GO_BP", lrrk2_results %>% filter(enrichment_type == "GO_BP"))

addWorksheet(wb, "KEGG")
writeData(wb, "KEGG", lrrk2_results %>% filter(enrichment_type == "KEGG"))

saveWorkbook(wb, "LRRK2_enrichment.xlsx", overwrite = TRUE)
```

## Troubleshooting

### Issue 1: "Dataset directory not valid"

**Symptoms**: Error message when launching app

**Solution**:
```{r troubleshoot-dataset}
# Check what's missing
validation <- validate_dataset_directory("/path/to/your/data/")

# View validation messages
print(validation$messages)

# Generate missing files if needed (the app offers to do this automatically)
```

### Issue 2: Out of Memory Errors

**Symptoms**: R crashes or "cannot allocate vector" errors

**Solutions**:

```{r troubleshoot-memory}
# 1. Increase memory limits
options(java.parameters = "-Xmx16g")  # 16GB

# 2. Use data sampling for exploration
sampled_data <- sample_de_results(
  de_results = full_results,
  sample_size = 50000  # Sample to 50k cells
)

# 3. Close other R sessions and applications

# 4. Process data in batches
```

### Issue 3: Slow App Performance

**Symptoms**: Slow plot rendering, lag in UI updates

**Solutions**:

```{r troubleshoot-performance}
# 1. The app uses caching - first load is slow, subsequent loads are fast
# Just wait for initial data load

# 2. Reduce data displayed:
#    - Filter to specific genes/clusters
#    - Use "Top N" terms instead of all terms in heatmaps

# 3. Use static plots instead of interactive for large heatmaps
```

### Issue 4: Cross-Platform Path Issues

**Symptoms**: "File not found" errors when moving between Windows/Mac/Linux

**Solutions**:

```{r troubleshoot-paths}
# Always use file.path() for cross-platform compatibility
data_dir <- file.path("path", "to", "data")  # Works on all platforms

# Reset configuration if needed
unlink(get_config_path())
launch_app()  # Will re-run setup wizard

# Check current configuration
get_parent_data_dir()
```

### Issue 5: "Module not found" Errors

**Symptoms**: Missing function or module errors

**Solutions**:

```{r troubleshoot-install}
# 1. Re-install package
remotes::install_github("jessedunnack/iSCORE-PDecipher", force = TRUE)

# 2. Check package version
packageVersion("iSCORE.PDecipher")

# 3. Restart R session
.rs.restartR()  # RStudio
# or quit() and restart R
```

## Next Steps

Now that you're familiar with the basics:

1. **Explore Other Vignettes**:
   - `vignette("perturbseq-workflow")`: Detailed Perturb-seq analysis
   - `vignette("signature-discovery")`: Advanced signature interpretation
   - `vignette("advanced-customization")`: Customization and batch processing

2. **Read the Documentation**:
   - Function reference: `?launch_app`, `?discover_top_signatures`, etc.
   - Package documentation: `help(package = "iSCORE.PDecipher")`

3. **Join the Community**:
   - Report bugs or request features: [GitHub Issues](https://github.com/jessedunnack/iSCORE-PDecipher/issues)
   - Contact: jessedunnack@berkeley.edu

## Session Info

```{r session-info}
sessionInfo()
```

## Citation

If you use iSCORE-PDecipher in your research, please cite:

```
Dunnack J. (2025). iSCORE-PDecipher: Integrated Analysis of
Parkinson's Disease Mutations and Perturbations.
GitHub: https://github.com/jessedunnack/iSCORE-PDecipher
```

---

**Package**: iSCORE.PDecipher
**Version**: 0.4.0+
**Author**: Jesse Dunnack
**License**: GPL-3
