---
title: "Perturb-seq Workflow: From Mixscale to Visualization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Perturb-seq Workflow: From Mixscale to Visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

This vignette provides a complete workflow for analyzing **CRISPRi Perturb-seq experiments** using iSCORE-PDecipher, with focus on the **pooled MixScale** approach and **FDR-corrected p-values**.

**What You'll Learn:**
- Import pooled MixScale results with FDR corrections
- Run enrichment analysis on Perturb-seq data
- Compare p-value correction methods (uncorrected vs BH vs Bonferroni)
- Visualize perturbation-specific effects
- Identify guide RNA efficiency
- Discover convergent pathways across perturbations

**Use Case**: You've run a CRISPRi screen targeting Parkinson's disease genes and want to understand which biological pathways are affected by each perturbation.

## Background: Pooled vs Experiment-Split Analysis

### Experiment-Split Approach (Original)

The original iSCORE-PDecipher was designed for **experiment-split** MixScale analysis:

- Separate analysis for each experiment (e.g., FPD-23, FPD-24, FPD-25)
- Column names like: `log2FC_C12_FPD-24`, `p_cell_typeC12_FPD-24:weight`
- Multiple columns per perturbation
- Useful when experiments have different conditions

### Pooled Approach (NEW)

**Pooled MixScale** analysis combines all experiments:

- Single analysis pooling all cells across experiments
- Simple column names: `log2FC`, `p_weight`, `p_weight_BH`, `p_weight_bonferroni`
- One value per gene per perturbation
- **Three p-value columns**:
  - `p_weight`: Original uncorrected p-values
  - `p_weight_BH`: Benjamini-Hochberg FDR correction (RECOMMENDED)
  - `p_weight_bonferroni`: Bonferroni correction (very conservative)
- More statistical power from larger sample size
- Control group: "Non-Targeting"

**When to use pooled:**
- Multiple batches/experiments with same conditions
- Want maximum statistical power
- Interested in overall perturbation effects

**When to use experiment-split:**
- Experiments have different treatment conditions
- Want to compare perturbation effects across conditions
- Investigating batch effects

## Example Dataset

This vignette uses the **all_CRISPRi dataset** with:

- **Perturbations**: 338 gene knockdowns (CRISPRi)
- **Clusters**: 6 cell clusters (cluster_0 through cluster_5)
- **Cells**: ~100,000 cells pooled across experiments
- **Control**: Non-Targeting guides
- **Target Genes**: ATP13A2, DNAJC6, FBXO7, LRRK2, PARK2, PARK7, PINK1, SNCA, SYNJ1, VPS13C, and many others

## Step 1: Import Pooled MixScale Data

### 1.1 Understanding the Data Structure

Pooled MixScale results are organized by cluster:

```
CRISPRi_PerturbSeq_Reports_all_CRISPRi_no_multiplets_noExptSplit/
├── all_CRISPRi_all_CRISPRi_no_multiplets_noExptSplit_Cluster0/
│   └── all_CRISPRi_no_multiplets_noExptSplit_clust_0_mixscale_DEGs.rds
├── all_CRISPRi_all_CRISPRi_no_multiplets_noExptSplit_Cluster1/
│   └── all_CRISPRi_no_multiplets_noExptSplit_clust_1_mixscale_DEGs.rds
...
└── all_CRISPRi_all_CRISPRi_no_multiplets_noExptSplit_Cluster5/
    └── all_CRISPRi_no_multiplets_noExptSplit_clust_5_mixscale_DEGs.rds
```

Each RDS file contains:
```r
# Structure: Named list of perturbations
list(
  "LRRK2" = data.frame(
    gene_ID = c("ENSG00000123", ...),
    log2FC = c(1.2, -0.8, ...),
    p_weight = c(0.001, 0.05, ...),              # Uncorrected
    p_weight_BH = c(0.01, 0.15, ...),            # BH FDR
    p_weight_bonferroni = c(0.02, 0.30, ...),    # Bonferroni
    beta_weight = c(0.5, -0.3, ...),
    ...
  ),
  "SNCA" = data.frame(...),
  ...
)
```

### 1.2 Import with iSCORE-PDecipher

```{r import-pooled}
library(iSCORE.PDecipher)

# Path to your pooled MixScale results
mixscale_dir <- "../final_hdWGCNA_results/testing_hdWGCNA/mixscale_for_paper/CRISPRi_PerturbSeq_Reports_all_CRISPRi_no_multiplets_noExptSplit/all_CRISPRi_all_CRISPRi_no_multiplets_noExptSplit_Cluster0/"

# Import with Benjamini-Hochberg FDR correction (recommended)
pooled_data_BH <- import_pooled_mixscale_data(
  mixscale_dir = mixscale_dir,
  pval_column = "p_weight_BH"  # Use FDR-corrected p-values
)

# Or import with uncorrected p-values for comparison
pooled_data_uncorrected <- import_pooled_mixscale_data(
  mixscale_dir = mixscale_dir,
  pval_column = "p_weight"
)

# Or import with Bonferroni correction (very conservative)
pooled_data_bonferroni <- import_pooled_mixscale_data(
  mixscale_dir = mixscale_dir,
  pval_column = "p_weight_bonferroni"
)
```

### 1.3 Verify Import

```{r verify-import}
# Check structure
str(pooled_data_BH, max.level = 2)

# How many perturbations?
cat("Number of perturbations:", length(pooled_data_BH), "\n")

# List perturbations
cat("Perturbations:\n")
print(names(pooled_data_BH))

# Check data for one perturbation
lrrk2_data <- pooled_data_BH$LRRK2
cat("\nLRRK2 perturbation results:\n")
cat("Genes tested:", nrow(lrrk2_data), "\n")
cat("Columns:", paste(names(lrrk2_data), collapse = ", "), "\n")

# Check for significant DEGs
sig_degs <- lrrk2_data[lrrk2_data$p_weight_BH < 0.05 & abs(lrrk2_data$log2FC) > 0.25, ]
cat("Significant DEGs (FDR < 0.05, |log2FC| > 0.25):", nrow(sig_degs), "\n")
```

## Step 2: Understanding P-Value Corrections

### 2.1 Why Multiple P-Value Corrections?

**Problem**: Multiple testing inflation
- Testing ~20,000 genes per perturbation
- 338 perturbations = ~6.7 million tests total
- Without correction, expect ~335,000 false positives at p < 0.05!

**Solutions**:

1. **Uncorrected (`p_weight`)**:
   - Raw p-values from weighted regression
   - Most liberal (highest sensitivity, lowest specificity)
   - Use for: Exploratory analysis, generating hypotheses

2. **Benjamini-Hochberg (`p_weight_BH`)** - RECOMMENDED:
   - Controls False Discovery Rate (FDR)
   - Balances sensitivity and specificity
   - GOLD STANDARD for scRNA-seq
   - Use for: Final results, publications

3. **Bonferroni (`p_weight_bonferroni`)**:
   - Controls Family-Wise Error Rate (FWER)
   - Most conservative (lowest sensitivity, highest specificity)
   - Use for: High-confidence hits, validation priorities

### 2.2 Compare Correction Methods

```{r compare-corrections}
# Load all three versions for same cluster
cluster0_uncorrected <- import_pooled_mixscale_data(mixscale_dir, pval_column = "p_weight")
cluster0_BH <- import_pooled_mixscale_data(mixscale_dir, pval_column = "p_weight_BH")
cluster0_bonferroni <- import_pooled_mixscale_data(mixscale_dir, pval_column = "p_weight_bonferroni")

# Compare significant DEG counts for LRRK2
threshold <- 0.05
lfc_threshold <- 0.25

count_sig_degs <- function(data, pert_name, pval_col, pval_thresh, lfc_thresh) {
  pert_data <- data[[pert_name]]
  sig <- pert_data[pert_data[[pval_col]] < pval_thresh &
                   abs(pert_data$log2FC) > lfc_thresh, ]
  nrow(sig)
}

# Count for LRRK2
cat("Significant DEGs for LRRK2 (p < 0.05, |log2FC| > 0.25):\n")
cat("Uncorrected:", count_sig_degs(cluster0_uncorrected, "LRRK2", "p_weight", threshold, lfc_threshold), "\n")
cat("BH FDR:", count_sig_degs(cluster0_BH, "LRRK2", "p_weight_BH", threshold, lfc_threshold), "\n")
cat("Bonferroni:", count_sig_degs(cluster0_bonferroni, "LRRK2", "p_weight_bonferroni", threshold, lfc_threshold), "\n")

# Expected pattern: Uncorrected > BH > Bonferroni
```

### 2.3 Visualize P-Value Distribution

```{r pval-distribution}
library(ggplot2)

# Extract LRRK2 p-values
lrrk2_data <- cluster0_BH$LRRK2

# Create data frame for plotting
pval_df <- data.frame(
  Uncorrected = lrrk2_data$p_weight,
  BH_FDR = lrrk2_data$p_weight_BH,
  Bonferroni = lrrk2_data$p_weight_bonferroni
)

# Reshape for ggplot
library(tidyr)
pval_long <- pivot_longer(pval_df, cols = everything(),
                          names_to = "Correction", values_to = "P_value")

# Plot distributions
ggplot(pval_long, aes(x = P_value, fill = Correction)) +
  geom_histogram(bins = 50, alpha = 0.6, position = "identity") +
  facet_wrap(~Correction, ncol = 1) +
  scale_x_log10() +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "red") +
  labs(title = "P-value Distributions: LRRK2 Perturbation",
       x = "P-value (log10 scale)",
       y = "Count") +
  theme_minimal()
```

## Step 3: Run Enrichment Analysis

### 3.1 Overview

Enrichment analysis identifies biological pathways and processes affected by each perturbation.

**Methods Available:**
- **GO (Gene Ontology)**: Biological Process (BP), Cellular Component (CC), Molecular Function (MF)
- **KEGG**: Metabolic and signaling pathways
- **Reactome**: Curated biological pathways
- **WikiPathways**: Community-curated pathways
- **STRING**: Protein-protein interaction networks
- **GSEA**: Gene Set Enrichment Analysis

### 3.2 Run Enrichment for All Perturbations

```{r run-enrichment}
# Path to cluster-specific results
cluster0_file <- "../final_hdWGCNA_results/testing_hdWGCNA/mixscale_for_paper/CRISPRi_PerturbSeq_Reports_all_CRISPRi_no_multiplets_noExptSplit/all_CRISPRi_all_CRISPRi_no_multiplets_noExptSplit_Cluster0/all_CRISPRi_no_multiplets_noExptSplit_clust_0_mixscale_DEGs.rds"

# Run complete enrichment analysis
enrichment_results <- run_enrichment_analysis(
  de_file = cluster0_file,
  output_dir = "./enrichment_output_cluster0/",
  pval_threshold = 0.05,
  lfc_threshold = 0.25,
  padj_method = "BH",  # Use BH FDR correction
  methods = c("GO_BP", "GO_MF", "GO_CC", "KEGG", "Reactome", "WikiPathways", "STRING", "GSEA")
)

# This will:
# 1. Load MixScale results
# 2. For each perturbation:
#    - Filter significant DEGs (p.adjust < 0.05, |log2FC| > 0.25)
#    - Separate UP and DOWN regulated genes
#    - Run enrichment for each method
#    - Save results in perturbation-specific directories
# 3. Return summary statistics

# Check summary
print(enrichment_results)
```

### 3.3 Import and Consolidate Enrichment Results

```{r import-enrichment}
# After enrichment analysis completes, import results for visualization

# Import enrichment with matching FDR correction
enrichment_data <- import_enrichment_with_correction(
  base_dir = "../final_hdWGCNA_results/testing_hdWGCNA/mixscale_for_paper/",
  dataset = "CRISPRi",
  pval_correction = "BH"  # Match correction used in DE analysis
)

# Consolidate for app
consolidated <- consolidate_pooled_enrichment(
  enrichment_dir = "./enrichment_output_cluster0/",
  output_file = "./cluster0_enrichment_consolidated.rds"
)

# Load consolidated results
enrichment_all <- readRDS("./cluster0_enrichment_consolidated.rds")

# Structure
cat("Enrichment terms:", nrow(enrichment_all), "\n")
cat("Columns:", paste(names(enrichment_all), collapse = ", "), "\n")
```

## Step 4: Visualize in Shiny App

### 4.1 Prepare Data for App

```{r prepare-for-app}
# Create directory structure for app
output_dir <- "./iscore_perturbseq_cluster0/"
dir.create(output_dir, recursive = TRUE)

# Copy/create required files:
# 1. full_DE_results.rds (converted from pooled format)
# 2. all_enrichment_padj005_complete_with_direction.rds (consolidated enrichment)

# Convert pooled MixScale to app format
converted_de <- convert_pooled_to_full_de(
  cluster_files = list(
    cluster_0 = cluster0_file
  ),
  output_file = file.path(output_dir, "full_DE_results.rds"),
  pval_column = "p_weight_BH"
)

# Copy consolidated enrichment
file.copy("./cluster0_enrichment_consolidated.rds",
          file.path(output_dir, "all_enrichment_padj005_complete_with_direction.rds"))
```

### 4.2 Launch Perturb-seq Focused App

```{r launch-app}
# Launch app with your Perturb-seq data
launch_app(data_dir = output_dir)

# The app will open in your browser
# Navigate through the interface to explore your results
```

### 4.3 Using the Perturb-seq Only Module

The app includes a specialized module for Perturb-seq data:

**Features:**
1. **P-value Correction Comparison**:
   - Radio buttons to switch between uncorrected, BH, and Bonferroni
   - Real-time updates to all visualizations

2. **Perturbation Table**:
   - List of all 338 perturbations
   - Number of DEGs per perturbation
   - Click to view details

3. **DEG Volcano Plot**:
   - Interactive volcano plot for selected perturbation
   - Hover to see gene names and statistics

4. **P-value Distribution**:
   - Histogram comparing uncorrected vs corrected p-values
   - Visualize FDR correction impact

5. **Enrichment Summary**:
   - Top enriched pathways for selected perturbation
   - Organized by enrichment type (GO_BP, KEGG, etc.)

**Workflow in App:**
```
1. Select perturbation (e.g., "LRRK2")
2. View DEG volcano plot
3. Check p-value distribution
4. Explore enrichment terms
5. Switch between p-value corrections to see impact
6. Export results
```

## Step 5: Perturbation-Specific Analysis

### 5.1 Identify High-Impact Perturbations

```{r high-impact}
library(dplyr)

# Count significant DEGs for each perturbation
deg_counts <- sapply(names(cluster0_BH), function(pert) {
  pert_data <- cluster0_BH[[pert]]
  sig <- pert_data[pert_data$p_weight_BH < 0.05 & abs(pert_data$log2FC) > 0.25, ]
  nrow(sig)
})

# Sort by DEG count
deg_counts_sorted <- sort(deg_counts, decreasing = TRUE)

# Top 20 perturbations by DEG count
cat("Top 20 Perturbations by Significant DEG Count:\n")
print(head(deg_counts_sorted, 20))

# Plot
barplot(head(deg_counts_sorted, 20),
        las = 2,  # Vertical labels
        cex.names = 0.7,
        main = "Top 20 Perturbations by DEG Count",
        ylab = "Number of Significant DEGs",
        col = "steelblue")
```

### 5.2 Analyze Specific Perturbation

Deep dive into LRRK2 perturbation:

```{r analyze-lrrk2}
# Extract LRRK2 data
lrrk2 <- cluster0_BH$LRRK2

# Significant DEGs
lrrk2_sig <- lrrk2[lrrk2$p_weight_BH < 0.05 & abs(lrrk2$log2FC) > 0.25, ]

# Separate UP and DOWN
lrrk2_up <- lrrk2_sig[lrrk2_sig$log2FC > 0, ]
lrrk2_down <- lrrk2_sig[lrrk2_sig$log2FC < 0, ]

cat("LRRK2 Perturbation Summary:\n")
cat("Total significant DEGs:", nrow(lrrk2_sig), "\n")
cat("Upregulated:", nrow(lrrk2_up), "\n")
cat("Downregulated:", nrow(lrrk2_down), "\n\n")

# Top upregulated genes
cat("Top 10 Upregulated Genes:\n")
lrrk2_up_sorted <- lrrk2_up[order(lrrk2_up$log2FC, decreasing = TRUE), ]
print(lrrk2_up_sorted[1:10, c("gene_ID", "log2FC", "p_weight_BH")])

# Top downregulated genes
cat("\nTop 10 Downregulated Genes:\n")
lrrk2_down_sorted <- lrrk2_down[order(lrrk2_down$log2FC), ]
print(lrrk2_down_sorted[1:10, c("gene_ID", "log2FC", "p_weight_BH")])

# Volcano plot
library(ggplot2)
lrrk2$significant <- ifelse(lrrk2$p_weight_BH < 0.05 & abs(lrrk2$log2FC) > 0.25,
                             "Significant", "Not Significant")

ggplot(lrrk2, aes(x = log2FC, y = -log10(p_weight_BH), color = significant)) +
  geom_point(alpha = 0.5) +
  scale_color_manual(values = c("Significant" = "red", "Not Significant" = "gray")) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  geom_vline(xintercept = c(-0.25, 0.25), linetype = "dashed") +
  labs(title = "LRRK2 Perturbation: Volcano Plot",
       x = "Log2 Fold Change",
       y = "-log10(FDR-adjusted p-value)") +
  theme_minimal()
```

### 5.3 Compare Related Perturbations

Compare LRRK2, SNCA, and GBA perturbations:

```{r compare-perturbations}
# Extract data for multiple perturbations
compare_perts <- c("LRRK2", "SNCA", "GBA")

comparison_list <- lapply(compare_perts, function(pert) {
  pert_data <- cluster0_BH[[pert]]
  sig <- pert_data[pert_data$p_weight_BH < 0.05 & abs(pert_data$log2FC) > 0.25, ]
  data.frame(
    Perturbation = pert,
    Total_DEGs = nrow(sig),
    Upregulated = sum(sig$log2FC > 0),
    Downregulated = sum(sig$log2FC < 0)
  )
})

comparison_df <- do.call(rbind, comparison_list)
print(comparison_df)

# Find shared DEGs
lrrk2_genes <- lrrk2_sig$gene_ID
snca_genes <- cluster0_BH$SNCA[cluster0_BH$SNCA$p_weight_BH < 0.05 &
                                abs(cluster0_BH$SNCA$log2FC) > 0.25, ]$gene_ID
gba_genes <- cluster0_BH$GBA[cluster0_BH$GBA$p_weight_BH < 0.05 &
                              abs(cluster0_BH$GBA$log2FC) > 0.25, ]$gene_ID

# Venn diagram (conceptually)
cat("\nShared DEGs:\n")
cat("LRRK2 & SNCA:", length(intersect(lrrk2_genes, snca_genes)), "\n")
cat("LRRK2 & GBA:", length(intersect(lrrk2_genes, gba_genes)), "\n")
cat("SNCA & GBA:", length(intersect(snca_genes, gba_genes)), "\n")
cat("All three:", length(Reduce(intersect, list(lrrk2_genes, snca_genes, gba_genes))), "\n")
```

## Step 6: Enrichment Analysis and Interpretation

### 6.1 Explore Enrichment for Specific Perturbation

```{r explore-enrichment}
# Load consolidated enrichment
enrichment_all <- readRDS("./cluster0_enrichment_consolidated.rds")

# Filter for LRRK2
lrrk2_enrichment <- enrichment_all %>%
  filter(gene == "LRRK2", cluster == "cluster_0", p.adjust < 0.05) %>%
  arrange(p.adjust)

# Top pathways by enrichment type
cat("Top 5 GO Biological Processes for LRRK2:\n")
lrrk2_go_bp <- lrrk2_enrichment %>%
  filter(enrichment_type == "GO_BP") %>%
  select(Description, GeneRatio, p.adjust, direction) %>%
  head(5)
print(lrrk2_go_bp)

cat("\nTop 5 KEGG Pathways for LRRK2:\n")
lrrk2_kegg <- lrrk2_enrichment %>%
  filter(enrichment_type == "KEGG") %>%
  select(Description, GeneRatio, p.adjust, direction) %>%
  head(5)
print(lrrk2_kegg)

# Count enriched terms by type and direction
enrichment_summary <- lrrk2_enrichment %>%
  group_by(enrichment_type, direction) %>%
  summarise(Count = n(), .groups = "drop")
print(enrichment_summary)
```

### 6.2 Visualize Enrichment

```{r visualize-enrichment}
# Dot plot of top GO_BP terms
library(ggplot2)

top_go_bp <- lrrk2_enrichment %>%
  filter(enrichment_type == "GO_BP") %>%
  arrange(p.adjust) %>%
  head(20)

# Parse GeneRatio to numeric
top_go_bp$GeneRatio_numeric <- sapply(strsplit(top_go_bp$GeneRatio, "/"),
                                       function(x) as.numeric(x[1]) / as.numeric(x[2]))

ggplot(top_go_bp, aes(x = GeneRatio_numeric, y = reorder(Description, GeneRatio_numeric))) +
  geom_point(aes(size = Count, color = p.adjust)) +
  scale_color_gradient(low = "red", high = "blue") +
  labs(title = "Top 20 GO Biological Processes: LRRK2 Perturbation",
       x = "Gene Ratio",
       y = "GO Term",
       color = "Adjusted p-value",
       size = "Gene Count") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 8))
```

### 6.3 Pathway Interpretation

Key biological themes for LRRK2 perturbation:

```{r interpret-pathways}
# Categorize pathways by biological theme
pd_relevant_terms <- c("mitochondria", "autophagy", "lysosome", "vesicle",
                       "endosome", "transport", "dopamine", "neuron",
                       "synaptic", "apoptosis", "protein", "ubiquitin")

lrrk2_enrichment$PD_relevant <- sapply(tolower(lrrk2_enrichment$Description), function(desc) {
  any(sapply(pd_relevant_terms, function(term) grepl(term, desc)))
})

cat("PD-relevant pathways for LRRK2:\n")
pd_pathways <- lrrk2_enrichment %>%
  filter(PD_relevant == TRUE, enrichment_type == "GO_BP") %>%
  arrange(p.adjust) %>%
  select(Description, GeneRatio, p.adjust, direction) %>%
  head(15)
print(pd_pathways)

# Categorize by biological process
categorize_pathway <- function(desc) {
  desc_lower <- tolower(desc)
  if (grepl("mitochondri", desc_lower)) return("Mitochondrial")
  if (grepl("autophagy|lysosom", desc_lower)) return("Autophagy/Lysosomal")
  if (grepl("vesicle|transport|endosom", desc_lower)) return("Vesicle Transport")
  if (grepl("dopamine|neuron|synaptic", desc_lower)) return("Neuronal Function")
  if (grepl("apoptosis|death", desc_lower)) return("Cell Death")
  if (grepl("protein|ubiquitin", desc_lower)) return("Protein Homeostasis")
  return("Other")
}

lrrk2_enrichment$Category <- sapply(lrrk2_enrichment$Description, categorize_pathway)

cat("\nPathway categories for LRRK2:\n")
category_counts <- table(lrrk2_enrichment$Category)
print(sort(category_counts, decreasing = TRUE))
```

## Step 7: Guide RNA Quality Control

### 7.1 Assessing Guide Efficiency

```{r guide-qc}
# Load cluster-specific MixScale results
cluster0_data <- readRDS(cluster0_file)

# For each perturbation, assess guide efficiency
# Guide efficiency can be inferred from:
# 1. Number of cells with guide
# 2. Effect size (log2FC)
# 3. Statistical significance

guide_efficiency <- data.frame(
  Perturbation = names(cluster0_data),
  Mean_AbsLog2FC = sapply(cluster0_data, function(x) mean(abs(x$log2FC), na.rm = TRUE)),
  N_Significant = sapply(cluster0_data, function(x) sum(x$p_weight_BH < 0.05 & abs(x$log2FC) > 0.25)),
  Max_Log2FC = sapply(cluster0_data, function(x) max(abs(x$log2FC), na.rm = TRUE))
)

# Sort by number of significant DEGs (proxy for efficiency)
guide_efficiency_sorted <- guide_efficiency[order(guide_efficiency$N_Significant, decreasing = TRUE), ]

cat("Top 10 Most Efficient Guides (by DEG count):\n")
print(head(guide_efficiency_sorted, 10))

cat("\nBottom 10 Least Efficient Guides:\n")
print(tail(guide_efficiency_sorted, 10))

# Plot efficiency distribution
hist(guide_efficiency$N_Significant,
     breaks = 50,
     main = "Distribution of Guide Efficiency",
     xlab = "Number of Significant DEGs",
     ylab = "Number of Perturbations",
     col = "steelblue")

# Identify low-efficiency guides (< 10 significant DEGs)
low_efficiency <- guide_efficiency[guide_efficiency$N_Significant < 10, ]
cat("\nLow efficiency guides (< 10 significant DEGs):", nrow(low_efficiency), "\n")
if (nrow(low_efficiency) > 0) {
  cat("Examples:\n")
  print(head(low_efficiency))
}
```

### 7.2 Effect Size vs Statistical Power

```{r effect-vs-power}
# Plot effect size vs number of significant DEGs
library(ggplot2)

ggplot(guide_efficiency, aes(x = Mean_AbsLog2FC, y = N_Significant)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_text(data = head(guide_efficiency_sorted, 10),
            aes(label = Perturbation),
            hjust = -0.1, vjust = -0.5, size = 3) +
  labs(title = "Guide Efficiency: Effect Size vs Statistical Power",
       x = "Mean Absolute Log2FC",
       y = "Number of Significant DEGs") +
  theme_minimal()

# Correlation between effect size and statistical power
cor_test <- cor.test(guide_efficiency$Mean_AbsLog2FC, guide_efficiency$N_Significant)
cat("\nCorrelation between effect size and DEG count:\n")
cat("r =", cor_test$estimate, ", p =", cor_test$p.value, "\n")
```

## Step 8: Cross-Cluster Analysis

### 8.1 Load Multiple Clusters

```{r load-multiple-clusters}
# Load data for all 6 clusters
clusters <- paste0("cluster_", 0:5)

cluster_dirs <- paste0(
  "../final_hdWGCNA_results/testing_hdWGCNA/mixscale_for_paper/CRISPRi_PerturbSeq_Reports_all_CRISPRi_no_multiplets_noExptSplit/all_CRISPRi_all_CRISPRi_no_multiplets_noExptSplit_Cluster",
  0:5, "/"
)

all_clusters_data <- lapply(1:6, function(i) {
  import_pooled_mixscale_data(
    mixscale_dir = cluster_dirs[i],
    pval_column = "p_weight_BH"
  )
})

names(all_clusters_data) <- clusters
```

### 8.2 Identify Cluster-Specific vs Pan-Cluster Effects

```{r cluster-comparison}
# For LRRK2, compare effects across clusters

lrrk2_across_clusters <- lapply(clusters, function(clust) {
  clust_data <- all_clusters_data[[clust]]
  lrrk2_data <- clust_data$LRRK2

  sig_degs <- lrrk2_data[lrrk2_data$p_weight_BH < 0.05 & abs(lrrk2_data$log2FC) > 0.25, ]

  data.frame(
    Cluster = clust,
    N_DEGs = nrow(sig_degs),
    N_Up = sum(sig_degs$log2FC > 0),
    N_Down = sum(sig_degs$log2FC < 0),
    Mean_Log2FC = mean(sig_degs$log2FC)
  )
})

lrrk2_cluster_summary <- do.call(rbind, lrrk2_across_clusters)
print(lrrk2_cluster_summary)

# Plot
barplot(lrrk2_cluster_summary$N_DEGs,
        names.arg = lrrk2_cluster_summary$Cluster,
        main = "LRRK2 Perturbation: DEGs Across Clusters",
        ylab = "Number of Significant DEGs",
        xlab = "Cluster",
        col = "steelblue")
```

### 8.3 Find Consistent Effects

```{r consistent-effects}
# Find genes consistently affected by LRRK2 across all clusters

# Extract significant genes from each cluster
lrrk2_genes_by_cluster <- lapply(all_clusters_data, function(clust_data) {
  lrrk2_data <- clust_data$LRRK2
  sig_degs <- lrrk2_data[lrrk2_data$p_weight_BH < 0.05 & abs(lrrk2_data$log2FC) > 0.25, ]
  sig_degs$gene_ID
})

# Find genes present in all clusters
consistent_genes <- Reduce(intersect, lrrk2_genes_by_cluster)
cat("Genes consistently affected by LRRK2 across ALL 6 clusters:", length(consistent_genes), "\n")
if (length(consistent_genes) > 0) {
  cat("Examples:\n")
  print(head(consistent_genes, 20))
}

# Find genes present in at least 4 clusters
gene_presence <- table(unlist(lrrk2_genes_by_cluster))
frequent_genes <- names(gene_presence[gene_presence >= 4])
cat("\nGenes affected in at least 4/6 clusters:", length(frequent_genes), "\n")
```

## Step 9: Export and Sharing Results

### 9.1 Export for Publication

```{r export-publication}
# Export LRRK2 results for manuscript

# 1. Significant DEGs to Excel
library(openxlsx)

wb <- createWorkbook()

# Sheet 1: All significant DEGs
addWorksheet(wb, "Significant_DEGs")
writeData(wb, "Significant_DEGs", lrrk2_sig)

# Sheet 2: Upregulated genes
addWorksheet(wb, "Upregulated")
writeData(wb, "Upregulated", lrrk2_up_sorted)

# Sheet 3: Downregulated genes
addWorksheet(wb, "Downregulated")
writeData(wb, "Downregulated", lrrk2_down_sorted)

# Sheet 4: Enrichment results
addWorksheet(wb, "Enrichment")
writeData(wb, "Enrichment", lrrk2_enrichment)

saveWorkbook(wb, "LRRK2_perturbation_results.xlsx", overwrite = TRUE)

# 2. Export volcano plot
pdf("LRRK2_volcano_plot.pdf", width = 8, height = 6)
# (Include ggplot code from above)
dev.off()

# 3. Export enrichment dot plot
pdf("LRRK2_enrichment_dotplot.pdf", width = 10, height = 8)
# (Include ggplot code from above)
dev.off()
```

### 9.2 Create Summary Report

```{r summary-report}
# Generate summary statistics for all perturbations

all_pert_summary <- data.frame(
  Perturbation = names(cluster0_BH),
  N_DEGs = sapply(cluster0_BH, function(x) sum(x$p_weight_BH < 0.05 & abs(x$log2FC) > 0.25)),
  N_Upregulated = sapply(cluster0_BH, function(x) sum(x$p_weight_BH < 0.05 & x$log2FC > 0.25)),
  N_Downregulated = sapply(cluster0_BH, function(x) sum(x$p_weight_BH < 0.05 & x$log2FC < -0.25)),
  Mean_Log2FC = sapply(cluster0_BH, function(x) mean(x$log2FC[x$p_weight_BH < 0.05], na.rm = TRUE))
)

# Sort by DEG count
all_pert_summary <- all_pert_summary[order(all_pert_summary$N_DEGs, decreasing = TRUE), ]

# Export to CSV
write.csv(all_pert_summary, "all_perturbations_summary_cluster0.csv", row.names = FALSE)

cat("Summary report saved to: all_perturbations_summary_cluster0.csv\n")
cat("Top 10 perturbations:\n")
print(head(all_pert_summary, 10))
```

## Best Practices and Tips

### 1. Always Use FDR Correction

For publication and final results, always use `p_weight_BH` (Benjamini-Hochberg):
```{r best-practice-1}
# GOOD
data <- import_pooled_mixscale_data(dir, pval_column = "p_weight_BH")

# USE WITH CAUTION (exploratory only)
data <- import_pooled_mixscale_data(dir, pval_column = "p_weight")
```

### 2. Set Reasonable Thresholds

```{r best-practice-2}
# Standard thresholds for Perturb-seq
pval_threshold <- 0.05    # FDR < 5%
lfc_threshold <- 0.25     # |log2FC| > 0.25 (~19% change)

# More stringent for high-confidence hits
pval_threshold <- 0.01    # FDR < 1%
lfc_threshold <- 0.5      # |log2FC| > 0.5 (~41% change)
```

### 3. Validate Key Findings

```{r best-practice-3}
# For important genes, validate with multiple approaches:
# 1. Check across multiple clusters
# 2. Verify with qPCR or Western blot
# 3. Compare with literature
# 4. Test with different FDR methods
```

### 4. Document Analysis Parameters

```{r best-practice-4}
# Save analysis metadata
analysis_metadata <- list(
  date = Sys.Date(),
  package_version = packageVersion("iSCORE.PDecipher"),
  dataset = "all_CRISPRi",
  clusters = "cluster_0",
  pval_correction = "BH",
  pval_threshold = 0.05,
  lfc_threshold = 0.25,
  enrichment_methods = c("GO_BP", "GO_MF", "GO_CC", "KEGG", "Reactome")
)

saveRDS(analysis_metadata, "analysis_metadata.rds")
```

## Troubleshooting

### Issue: "No significant DEGs found"

**Possible causes:**
1. Thresholds too stringent
2. Poor guide efficiency
3. Low statistical power (too few cells)

**Solutions:**
```{r troubleshoot-no-degs}
# 1. Relax thresholds
pval_threshold <- 0.1
lfc_threshold <- 0.2

# 2. Check guide efficiency
# (See Section 7.1)

# 3. Use uncorrected p-values for exploration
data_uncorrected <- import_pooled_mixscale_data(dir, pval_column = "p_weight")
```

### Issue: "Too many DEGs"

**Possible causes:**
1. Using uncorrected p-values
2. Very strong perturbation effect

**Solutions:**
```{r troubleshoot-too-many}
# 1. Use FDR correction
data_BH <- import_pooled_mixscale_data(dir, pval_column = "p_weight_BH")

# 2. Increase LFC threshold
lfc_threshold <- 0.5  # More stringent

# 3. Use Bonferroni for highest confidence
data_bonferroni <- import_pooled_mixscale_data(dir, pval_column = "p_weight_bonferroni")
```

## Conclusions

This workflow demonstrates:

1. ✅ Importing pooled MixScale data with FDR corrections
2. ✅ Comparing p-value correction methods
3. ✅ Running comprehensive enrichment analysis
4. ✅ Visualizing results in Shiny app
5. ✅ Perturbation-specific analysis
6. ✅ Guide RNA quality control
7. ✅ Cross-cluster comparisons
8. ✅ Exporting publication-ready results

**Key Takeaways:**
- Always use FDR correction (BH recommended) for final results
- Pooled analysis provides maximum statistical power
- iSCORE-PDecipher makes Perturb-seq analysis accessible and visual
- Cross-cluster analysis identifies robust effects
- Guide QC is essential for data quality

## Next Steps

- Explore **Signature Discovery** vignette for cross-method analysis
- Read **Advanced Customization** vignette for batch processing
- Use the **Getting Started** vignette for app basics

## Session Info

```{r session-info}
sessionInfo()
```
