---
title: "Signature Discovery Tutorial: Finding Convergent Pathways"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Signature Discovery Tutorial: Finding Convergent Pathways}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

**Signature Discovery** is one of the most powerful features of iSCORE-PDecipher, allowing you to identify **convergent pathways** - biological processes affected by BOTH a genetic mutation (MAST) AND a gene knockdown perturbation (CRISPRi).

**What are Signatures?**

A **signature** is a pathway or biological process that shows:
1. Enrichment in genes affected by a genetic mutation (e.g., LRRK2_G2019S)
2. Enrichment in genes affected by a perturbation (e.g., LRRK2 CRISPRi knockdown)
3. Statistical evidence of gene overlap between the two conditions
4. Correlated effect sizes (similar direction and magnitude of changes)

**Why are Signatures Important?**

- **Cross-validation**: Findings validated by two independent methods (MAST + MixScale)
- **Robustness**: Less likely to be false positives or technical artifacts
- **Biological insight**: Reveal core pathways affected by gene dysfunction
- **Therapeutic relevance**: Convergent pathways are strong targets for intervention

**This Vignette Covers:**
1. Understanding signature scoring algorithms
2. Discovering signatures programmatically
3. Using the Signature Nomination Shiny module
4. Interpreting PD-relevant signatures
5. Pan-cluster vs cluster-specific signatures
6. Advanced signature analysis workflows

## Background: The Signature Discovery Algorithm

### Step 1: Enrichment Term Collection

For a given gene pair (e.g., LRRK2 mutation + LRRK2 perturbation):

```
MAST Results:                    MixScale Results:
Cluster 0, LRRK2_G2019S         Cluster 0, LRRK2 CRISPRi
↓                               ↓
Enrichment Analysis             Enrichment Analysis
↓                               ↓
Enriched Terms (MAST)          Enriched Terms (MixScale)
- mitochondrial transport       - mitochondrial transport
- autophagy                     - vesicle transport
- lysosomal degradation         - lysosomal degradation
- protein phosphorylation       - autophagy
...                            ...
```

### Step 2: Gene Overlap Analysis

For each pathway found in both methods:

```r
# Example: "mitochondrial transport"
MAST_genes <- c("TOMM20", "TIMM23", "TOMM40", "MFN2", "OPA1", ...)  # 25 genes
MixScale_genes <- c("TOMM20", "TIMM23", "TOMM40", "MFN1", ...)      # 20 genes
Overlap_genes <- c("TOMM20", "TIMM23", "TOMM40", ...)               # 15 genes

# Fisher's Exact Test
# H0: Gene overlap is random
# H1: Gene overlap is significant

contingency_table <- matrix(c(
  overlap_count,        mast_only_count,
  mixscale_only_count,  neither_count
), nrow = 2)

fisher_result <- fisher.test(contingency_table)
# p-value: < 0.001 → Significant overlap!
```

### Step 3: Effect Size Correlation

Compare log2 fold changes for overlapping genes:

```r
# For genes in both MAST and MixScale
mast_log2fc     <- c(1.2, -0.8, 1.5, -1.1, ...)
mixscale_log2fc <- c(1.0, -0.9, 1.3, -1.0, ...)

correlation <- cor.test(mast_log2fc, mixscale_log2fc)
# r = 0.85, p < 0.001 → Strong positive correlation!
```

### Step 4: Composite Signature Score

Combine multiple metrics into a signature score:

```r
signature_score <- weighted_average(
  gene_overlap_score,      # Fisher's p-value
  effect_correlation,      # Pearson's r
  direction_consistency,   # % genes with same direction
  enrichment_strength      # Combined enrichment p-values
)

# Score ranges from 0 (weak) to 1 (strong)
# Threshold: 0.7 for high-confidence signatures
```

## Part 1: Programmatic Signature Discovery

### 1.1 Load Required Data

```{r load-data}
library(iSCORE.PDecipher)
library(dplyr)

# Load consolidated enrichment data
enrichment_data <- readRDS("all_enrichment_padj005_complete_with_direction.rds")

# Check structure
cat("Total enrichment terms:", nrow(enrichment_data), "\n")
cat("Columns:", paste(names(enrichment_data), collapse = ", "), "\n")
cat("Methods:", unique(enrichment_data$method), "\n")
cat("Enrichment types:", unique(enrichment_data$enrichment_type), "\n")
```

### 1.2 Discover Top Signatures

```{r discover-signatures}
# Run signature discovery for all gene pairs
signatures <- discover_top_signatures(
  enrichment_data = enrichment_data,
  top_n = 50,                    # Return top 50 signatures
  min_cluster_presence = 3,      # Must appear in at least 3 clusters
  fisher_pval_threshold = 0.01,  # Fisher's test p < 0.01
  fdr_correction = TRUE,         # Apply FDR correction
  min_overlap_genes = 3          # At least 3 overlapping genes
)

# Results structure
cat("Signatures discovered:", nrow(signatures), "\n")
cat("Columns:", paste(names(signatures), collapse = ", "), "\n")

# Top 10 signatures
head(signatures, 10)
```

### 1.3 Explore Signature Results

```{r explore-signatures}
# Sort by signature score
signatures_sorted <- signatures[order(signatures$signature_score, decreasing = TRUE), ]

# Examine top signature
top_sig <- signatures_sorted[1, ]
cat("Top Signature:\n")
cat("MAST Gene:", top_sig$mast_gene, "\n")
cat("MixScale Gene:", top_sig$mixscale_gene, "\n")
cat("Cluster:", top_sig$cluster, "\n")
cat("Pathway:", top_sig$pathway_description, "\n")
cat("Signature Score:", top_sig$signature_score, "\n")
cat("Fisher p-value:", top_sig$fisher_pvalue, "\n")
cat("Effect Correlation:", top_sig$effect_correlation, "\n")
cat("Overlap Genes:", top_sig$n_overlap_genes, "\n")

# Gene list in signature
cat("\nGenes in signature:\n")
overlap_genes <- strsplit(top_sig$overlap_gene_ids, "/")[[1]]
print(overlap_genes)
```

### 1.4 Filter and Analyze Signatures

```{r filter-signatures}
# Filter for high-confidence signatures
high_confidence <- signatures %>%
  filter(
    signature_score > 0.7,
    fisher_pvalue < 0.001,
    n_overlap_genes >= 5
  )

cat("High-confidence signatures:", nrow(high_confidence), "\n")

# Group by pathway to find frequently converging processes
pathway_frequency <- high_confidence %>%
  group_by(pathway_description) %>%
  summarise(
    Count = n(),
    Mean_Score = mean(signature_score),
    Gene_Pairs = paste(unique(paste(mast_gene, mixscale_gene, sep = "+")), collapse = ", ")
  ) %>%
  arrange(desc(Count))

cat("\nTop 10 Most Frequent Converging Pathways:\n")
print(head(pathway_frequency, 10))

# Find LRRK2-specific signatures
lrrk2_signatures <- signatures %>%
  filter(mast_gene == "LRRK2_G2019S" | mixscale_gene == "LRRK2") %>%
  arrange(desc(signature_score))

cat("\nLRRK2-specific signatures:", nrow(lrrk2_signatures), "\n")
head(lrrk2_signatures[, c("mast_gene", "mixscale_gene", "pathway_description", "signature_score")], 10)
```

## Part 2: Using the Signature Nomination Shiny Module

### 2.1 Launch App

```{r launch-app}
# Launch iSCORE-PDecipher
launch_app()

# Navigate to "Signature Nomination" tab in the app
```

### 2.2 Tab 1: Signature Discovery

**Step-by-Step Workflow:**

```
1. Select Gene Pair:
   - MAST Gene (Mutation): "LRRK2_G2019S"
   - MixScale Gene (Perturbation): "LRRK2"

2. Click "Discover Signatures" button

3. Review Results Table:
   Columns:
   - Pathway: Name of converging pathway
   - Overlap Genes: Number and list of genes
   - Fisher p-value: Statistical significance
   - Effect Correlation: Pearson's r for log2FC
   - Signature Score: Composite metric (0-1)

4. Sort by Signature Score (highest first)

5. Click on pathway row to see:
   - Scatter plot: MAST log2FC vs MixScale log2FC
   - Venn diagram: Gene overlap visualization
   - Gene table: Details for each overlapping gene
```

**Interactive Features:**

- **Scatter Plot**:
  - X-axis: MAST log2 fold change
  - Y-axis: MixScale log2 fold change
  - Each point = one gene in pathway
  - Color: Direction consistency (same/opposite)
  - Hover: See gene name and values

- **Venn Diagram**:
  - Left circle: MAST-enriched genes
  - Right circle: MixScale-enriched genes
  - Overlap: Convergent genes (signature)
  - Numbers: Gene counts in each region

- **Advanced Options** (sidebar):
  - Fisher p-value threshold: Adjust stringency
  - Minimum overlap genes: Filter weak signatures
  - FDR correction method: BH, Bonferroni, or none
  - Cluster filter: Analyze specific clusters

### 2.3 Tab 2: PD Biology Focus

This tab provides **biological interpretation** of signatures with Parkinson's disease relevance.

**Features:**

1. **Biological Categories**:
   - Mitochondrial function
   - Autophagy/Lysosomal degradation
   - Dopamine signaling
   - Protein aggregation
   - Neuroinflammation
   - Synaptic function
   - ER stress/UPR
   - Oxidative stress

2. **PD Relevance Scoring**:
   - Automatic scoring based on PD literature keywords
   - Higher scores = more relevant to PD pathology

3. **Pathway Interpretation**:
   - AI-generated biological context
   - Explanation of pathway relevance to PD
   - Literature references

**Workflow:**

```
1. Select biological categories of interest:
   ☑ Mitochondrial function
   ☑ Autophagy/Lysosomal
   ☐ Dopamine signaling
   ...

2. Click "Filter by PD Biology"

3. Review filtered signatures:
   - Only signatures in selected categories shown
   - Sorted by PD relevance score

4. Click on pathway for interpretation:
   - Biological context paragraph
   - PD relevance explanation
   - Related signatures

5. Category Heatmap:
   - Rows: Gene pairs
   - Columns: Biological categories
   - Colors: Number of signatures in each category
   - Identify gene pairs with broad pathway disruption
```

**Example Interpretation:**

```
Pathway: "mitochondrial transport"
PD Relevance Score: 0.92

Interpretation:
Mitochondrial transport is critically impaired in Parkinson's disease. The PINK1-Parkin
pathway, which regulates mitophagy (selective removal of damaged mitochondria), is directly
affected by mutations in PARK2 (Parkin) and PINK1. This signature shows convergent disruption
of mitochondrial protein import (TOMM20, TIMM23) and fusion/fission dynamics (MFN2, OPA1),
suggesting a core vulnerability in dopaminergic neurons.

Literature: Narendra et al., 2008; Pickrell & Youle, 2015
```

### 2.4 Tab 3: Pan-Cluster Signatures

**Purpose**: Identify signatures that appear consistently across multiple cell clusters.

**Rationale**: Signatures present in many clusters represent **robust, cell-type-independent effects** - the strongest evidence for convergence.

**Workflow:**

```
1. Adjust "Minimum Clusters" slider:
   - Set to 5 (must appear in 5+ clusters)

2. Review Pan-Cluster Table:
   - Pathway: Converging process
   - Clusters Present: Which clusters show signature
   - Mean Signature Score: Average across clusters
   - Gene Pairs: Which gene pairs show this signature

3. Click pathway to see:
   - Cluster Presence Heatmap:
     * Rows: Gene pairs
     * Columns: Clusters (cluster_0 to cluster_14)
     * Colors: Signature score in each cluster
     * White: Signature absent
     * Red: Strong signature

4. Identify "Universal Signatures":
   - Present in ALL clusters (or most)
   - High scores across clusters
   - Core pathways affected by gene dysfunction
```

**Example Pan-Cluster Signatures:**

```r
# Programmatically find pan-cluster signatures
pan_cluster_sigs <- signatures %>%
  group_by(pathway_description, mast_gene, mixscale_gene) %>%
  summarise(
    N_Clusters = n_distinct(cluster),
    Mean_Score = mean(signature_score),
    Clusters = paste(unique(cluster), collapse = ", ")
  ) %>%
  filter(N_Clusters >= 5) %>%
  arrange(desc(Mean_Score))

cat("Top 5 Pan-Cluster Signatures:\n")
print(head(pan_cluster_sigs, 5))

# Expected results (hypothetical):
#   pathway_description              mast_gene     mixscale_gene  N_Clusters  Mean_Score
#   mitochondrial transport          LRRK2_G2019S  LRRK2          8           0.85
#   autophagosome assembly           PARK2         PARK2          7           0.82
#   protein ubiquitination           LRRK2_G2019S  LRRK2          6           0.79
#   lysosomal protein catabolic...   GBA           GBA            6           0.77
#   vesicle-mediated transport       SNCA_A53T     SNCA           5           0.74
```

### 2.5 Tab 4: Cluster-Specific Signatures

**Purpose**: Identify signatures unique to individual cell clusters.

**Rationale**: Cluster-specific signatures reveal **cell-type-specific vulnerabilities** relevant to PD heterogeneity.

**Workflow:**

```
1. Select cluster: "cluster_0" (e.g., dopaminergic neurons)

2. Review Cluster-Specific Table:
   - Signatures unique to this cluster
   - Or much stronger in this cluster vs others

3. Compare across clusters:
   - Select different clusters
   - Note which pathways are cluster-specific
   - Identify cell-type vulnerabilities

4. Export cluster-specific signatures:
   - Download table for manuscript
   - Include in cell-type characterization
```

**Example**:
- Cluster 0 (dopaminergic): Dopamine metabolism signatures
- Cluster 5 (astrocytes): Glial support function signatures
- Cluster 10 (microglia): Neuroinflammation signatures

## Part 3: Advanced Signature Analysis

### 3.1 Biological Context Extraction

```{r extract-context}
# For a specific signature, extract full biological context
sig_context <- extract_signature_biological_context(
  mast_gene = "LRRK2_G2019S",
  mixscale_gene = "LRRK2",
  cluster = "cluster_0",
  enrichment_data = enrichment_data
)

# Results include:
cat("Signature context:\n")
cat("Shared pathways:", length(sig_context$shared_pathways), "\n")
cat("MAST-only pathways:", length(sig_context$mast_only), "\n")
cat("MixScale-only pathways:", length(sig_context$mixscale_only), "\n")
cat("PD relevance score:", sig_context$pd_relevance_score, "\n")

# View shared pathways
cat("\nTop shared pathways:\n")
print(head(sig_context$shared_pathways, 10))

# Categorize by biological process
categorized <- categorize_biological_processes(sig_context$shared_pathways)
cat("\nBiological process distribution:\n")
print(table(categorized$category))
```

### 3.2 Direction Consistency Analysis

```{r direction-consistency}
# Analyze whether genes change in same direction

direction_analysis <- calculate_direction_consistency(
  mast_gene = "LRRK2_G2019S",
  mixscale_gene = "LRRK2",
  cluster = "cluster_0",
  enrichment_data = enrichment_data
)

cat("Direction consistency metrics:\n")
cat("% Same direction:", direction_analysis$pct_same_direction, "\n")
cat("% Opposite direction:", direction_analysis$pct_opposite_direction, "\n")
cat("Direction concordance score:", direction_analysis$concordance_score, "\n")

# Interpretation:
# High concordance (>0.8): Strong convergence
# Medium concordance (0.5-0.8): Partial convergence
# Low concordance (<0.5): Weak or antagonistic effects
```

### 3.3 Signature Trends Analysis

Discover signatures without manual curation:

```{r signature-trends}
# Data-driven signature discovery
trends <- analyze_signature_trends(
  enrichment_data = enrichment_data,
  top_n = 100
)

# Three analysis types:

# 1. Frequency Analysis: Most common signatures
frequent_sigs <- trends$frequency_analysis
cat("Top 10 Most Frequent Signatures:\n")
print(head(frequent_sigs, 10))

# 2. Impact Analysis: Strongest effect sizes
impact_sigs <- trends$impact_analysis
cat("\nTop 10 Highest Impact Signatures:\n")
print(head(impact_sigs, 10))

# 3. Term Pattern Analysis: Recurring enrichment patterns
patterns <- trends$term_patterns
cat("\nEnrichment term patterns discovered:\n")
cat("Ubiquitous terms:", length(patterns$ubiquitous), "\n")
cat("Cluster-specific terms:", length(patterns$cluster_specific), "\n")
cat("Gene-specific terms:", length(patterns$gene_specific), "\n")
```

### 3.4 Multi-Gene Signature Discovery

Find signatures involving multiple gene pairs:

```{r multi-gene-signatures}
# Define PD gene sets
pd_mutations <- c("LRRK2_G2019S", "SNCA_A53T", "GBA", "PARK7")
pd_perturbations <- c("LRRK2", "SNCA", "GBA", "PARK7")

# Discover signatures for all combinations
all_pd_signatures <- lapply(pd_mutations, function(mut) {
  lapply(pd_perturbations, function(pert) {
    discover_top_signatures(
      enrichment_data = enrichment_data,
      mast_gene_filter = mut,
      mixscale_gene_filter = pert,
      top_n = 20
    )
  })
})

# Flatten results
all_sigs_combined <- do.call(rbind, lapply(all_pd_signatures, function(x) do.call(rbind, x)))

# Find pathways converging across multiple gene pairs
pathway_convergence <- all_sigs_combined %>%
  group_by(pathway_description) %>%
  summarise(
    N_Gene_Pairs = n_distinct(paste(mast_gene, mixscale_gene)),
    Mean_Score = mean(signature_score),
    Gene_Pairs = paste(unique(paste(mast_gene, mixscale_gene, sep = "+")), collapse = "; ")
  ) %>%
  filter(N_Gene_Pairs >= 3) %>%
  arrange(desc(N_Gene_Pairs))

cat("Pathways converging across multiple PD genes:\n")
print(pathway_convergence)

# These are "core PD pathways" - affected by many PD genes
```

## Part 4: Visualization and Export

### 4.1 Create Signature Heatmaps

```{r signature-heatmap}
# Interactive signature heatmap
library(heatmaply)

# Prepare data matrix
# Rows: Pathways, Columns: Gene pairs, Values: Signature scores
signature_matrix <- signatures %>%
  mutate(GenePair = paste(mast_gene, mixscale_gene, sep = "_")) %>%
  select(pathway_description, GenePair, signature_score) %>%
  pivot_wider(names_from = GenePair, values_from = signature_score, values_fill = 0) %>%
  column_to_rownames("pathway_description") %>%
  as.matrix()

# Create interactive heatmap
heatmaply(
  signature_matrix,
  scale = "none",
  colors = colorRampPalette(c("white", "yellow", "red"))(100),
  main = "Signature Convergence Across Gene Pairs",
  xlab = "Gene Pair (Mutation + Perturbation)",
  ylab = "Converging Pathway",
  fontsize_row = 8,
  fontsize_col = 8
)

# Export as HTML
htmlwidgets::saveWidget(
  heatmaply(signature_matrix),
  "signature_heatmap_interactive.html"
)
```

### 4.2 Create Signature Network

```{r signature-network}
# Network visualization showing gene-pathway relationships
library(igraph)
library(ggraph)

# Prepare edge list
edges <- signatures %>%
  filter(signature_score > 0.7) %>%
  select(
    from = mast_gene,
    to = pathway_description,
    weight = signature_score
  )

# Create graph
sig_network <- graph_from_data_frame(edges, directed = FALSE)

# Plot
ggraph(sig_network, layout = "fr") +
  geom_edge_link(aes(alpha = weight), color = "gray") +
  geom_node_point(aes(color = degree(sig_network)), size = 3) +
  geom_node_text(aes(label = name), size = 3, repel = TRUE) +
  scale_color_gradient(low = "lightblue", high = "darkred") +
  labs(title = "Signature Network: Genes and Converging Pathways") +
  theme_void()

# Save
ggsave("signature_network.pdf", width = 12, height = 10)
```

### 4.3 Exporting Results

```{r export-signatures}
# Export high-confidence signatures to Excel

library(openxlsx)

wb <- createWorkbook()

# Sheet 1: All signatures
addWorksheet(wb, "All_Signatures")
writeData(wb, "All_Signatures", signatures)

# Sheet 2: High-confidence only
high_conf <- signatures %>% filter(signature_score > 0.7)
addWorksheet(wb, "High_Confidence")
writeData(wb, "High_Confidence", high_conf)

# Sheet 3: Pan-cluster signatures
addWorksheet(wb, "Pan_Cluster")
writeData(wb, "Pan_Cluster", pan_cluster_sigs)

# Sheet 4: Pathway frequency
addWorksheet(wb, "Pathway_Frequency")
writeData(wb, "Pathway_Frequency", pathway_frequency)

# Sheet 5: PD-relevant signatures
pd_relevant <- signatures %>%
  filter(grepl("mitochondri|autophagy|lysosom|dopamin|synap", pathway_description, ignore.case = TRUE))
addWorksheet(wb, "PD_Relevant")
writeData(wb, "PD_Relevant", pd_relevant)

# Save workbook
saveWorkbook(wb, "signature_discovery_results.xlsx", overwrite = TRUE)

cat("Results exported to: signature_discovery_results.xlsx\n")
```

## Part 5: Biological Interpretation

### 5.1 Interpreting Signature Scores

**Score Range Interpretation:**

| Score Range | Interpretation | Confidence | Recommendation |
|-------------|----------------|------------|----------------|
| 0.9 - 1.0 | Extremely strong convergence | Very High | Priority for validation |
| 0.7 - 0.9 | Strong convergence | High | Include in manuscript |
| 0.5 - 0.7 | Moderate convergence | Medium | Exploratory analysis |
| 0.3 - 0.5 | Weak convergence | Low | Further investigation needed |
| 0.0 - 0.3 | Minimal convergence | Very Low | Likely false positive |

**Score Components:**

1. **Gene Overlap** (30% weight):
   - Fisher's exact test p-value
   - Number of overlapping genes
   - Lower p-value = higher score

2. **Effect Correlation** (30% weight):
   - Pearson's r for log2FC values
   - Higher correlation = higher score
   - Positive correlation expected for true convergence

3. **Direction Consistency** (20% weight):
   - % of genes changing in same direction
   - Same direction = convergent disruption
   - Opposite direction = antagonistic effects

4. **Enrichment Strength** (20% weight):
   - Combined p-values from both methods
   - Stronger enrichment = higher score

### 5.2 PD-Relevant Pathway Categories

**Category 1: Mitochondrial Function**

Pathways:
- Mitochondrial transport
- Oxidative phosphorylation
- Mitochondrial dynamics (fusion/fission)
- Calcium homeostasis

PD Relevance:
- Mitochondrial dysfunction is a hallmark of PD
- PINK1-Parkin pathway directly regulates mitophagy
- Complex I deficiency common in PD patients

Example Signature:
```
Pathway: "mitochondrial protein import"
Gene Pair: LRRK2_G2019S + LRRK2
Signature Score: 0.87
Overlap Genes: TOMM20, TIMM23, TOMM40, SAMM50, MFN2
Interpretation: LRRK2 disrupts mitochondrial protein import,
  leading to accumulation of misfolded proteins in mitochondria
```

**Category 2: Autophagy/Lysosomal Degradation**

Pathways:
- Autophagosome assembly
- Lysosomal protein catabolism
- Endosome-lysosome fusion
- LC3 lipidation

PD Relevance:
- GBA mutations (glucocerebrosidase) directly affect lysosomal function
- Impaired autophagy leads to α-synuclein accumulation
- Lewy bodies contain autophagy markers

Example Signature:
```
Pathway: "autophagosome-lysosome fusion"
Gene Pair: GBA + GBA
Signature Score: 0.91
Overlap Genes: ATG5, ATG7, LC3B, LAMP1, LAMP2
Interpretation: GBA loss-of-function impairs autophagy completion,
  preventing clearance of protein aggregates
```

**Category 3: Vesicle Transport**

Pathways:
- Vesicle-mediated transport
- Golgi-ER trafficking
- Exocytosis
- Synaptic vesicle recycling

PD Relevance:
- α-synuclein plays a role in vesicle trafficking
- VPS35 mutations affect retromer complex
- Dopamine release depends on vesicle fusion

**Category 4: Protein Homeostasis**

Pathways:
- Protein ubiquitination
- Proteasome-mediated degradation
- Unfolded protein response
- Chaperone-mediated autophagy

PD Relevance:
- PARK2 encodes E3 ubiquitin ligase
- Proteasome impairment in PD
- α-synuclein aggregation from misfolding

**Category 5: Dopamine Metabolism**

Pathways:
- Dopamine biosynthesis
- Catecholamine transport
- Monoamine oxidase activity
- Dopamine receptor signaling

PD Relevance:
- Loss of dopaminergic neurons defines PD
- Dopamine oxidation products are toxic
- L-DOPA therapy targets this pathway

### 5.3 Cross-Method Validation

**Why Convergence Matters:**

```
Single Method (MAST only):
✗ Could be mutation-specific artifact
✗ Could be experimental bias
✗ Limited biological confidence

Single Method (MixScale only):
✗ Could be off-target effect
✗ Could be perturbation artifact
✗ Guide RNA efficiency issues

Convergent (MAST + MixScale):
✓ Validated by two independent methods
✓ Robust to technical artifacts
✓ High biological confidence
✓ Strong therapeutic target
```

**Interpretation Guidelines:**

1. **High-confidence signatures (score > 0.7)**:
   - Include in manuscript main figures
   - Priority for experimental validation
   - Consider for therapeutic targeting

2. **Medium-confidence signatures (score 0.5-0.7)**:
   - Include in supplementary figures
   - Validate with additional assays
   - Exploratory analysis

3. **Low-confidence signatures (score < 0.5)**:
   - Hypothesis generation only
   - Require extensive validation
   - May be false positives

## Part 6: Case Studies

### Case Study 1: LRRK2_G2019S Signatures

**Question**: What pathways converge between LRRK2 G2019S mutation and LRRK2 CRISPRi knockdown?

```{r case-study-lrrk2}
# Discover LRRK2 signatures
lrrk2_sigs <- discover_top_signatures(
  enrichment_data = enrichment_data,
  mast_gene_filter = "LRRK2_G2019S",
  mixscale_gene_filter = "LRRK2",
  top_n = 50,
  min_cluster_presence = 1
)

# Analyze by cluster
lrrk2_by_cluster <- lrrk2_sigs %>%
  group_by(cluster) %>%
  summarise(
    N_Signatures = n(),
    Top_Pathway = pathway_description[which.max(signature_score)],
    Top_Score = max(signature_score)
  )

print(lrrk2_by_cluster)

# Key findings (hypothetical):
# cluster_0: Mitochondrial transport (score: 0.89)
# cluster_1: Vesicle-mediated transport (score: 0.85)
# cluster_2: Autophagosome assembly (score: 0.82)
```

**Interpretation**:
- LRRK2 disrupts multiple pathways across cell types
- Mitochondrial dysfunction is strongest signature
- Consistent with LRRK2's role in mitochondrial dynamics
- Validates therapeutic strategies targeting mitochondria

### Case Study 2: Pan-PD Gene Signatures

**Question**: Which pathways are affected by multiple PD genes?

```{r case-study-pan-pd}
# Core PD genes
pd_genes <- c("LRRK2", "SNCA", "GBA", "PARK2", "PINK1", "PARK7")

# Find signatures for each gene (mutation + perturbation)
pan_pd_sigs <- lapply(pd_genes, function(gene) {
  discover_top_signatures(
    enrichment_data = enrichment_data,
    mast_gene_filter = paste0(gene, "_"),  # Match any mutation variant
    mixscale_gene_filter = gene,
    top_n = 30
  )
})

pan_pd_combined <- do.call(rbind, pan_pd_sigs)

# Find pathways appearing in multiple genes
core_pathways <- pan_pd_combined %>%
  group_by(pathway_description) %>%
  summarise(
    N_Genes = n_distinct(mast_gene),
    Mean_Score = mean(signature_score),
    Genes = paste(unique(mast_gene), collapse = ", ")
  ) %>%
  filter(N_Genes >= 3) %>%
  arrange(desc(N_Genes))

print(core_pathways)

# Expected core pathways:
# 1. Mitochondrial transport (6 genes)
# 2. Autophagy (5 genes)
# 3. Protein ubiquitination (5 genes)
# 4. Lysosomal degradation (4 genes)
# 5. Vesicle transport (4 genes)
```

**Interpretation**:
- Core pathways represent shared PD mechanisms
- Therapeutic targets should address these convergent processes
- Multiple gene mutations converge on same cellular dysfunction

## Best Practices

### 1. Always Apply FDR Correction

```{r best-practice-fdr}
# GOOD: Use FDR-corrected signatures
signatures <- discover_top_signatures(
  enrichment_data,
  fdr_correction = TRUE,
  fisher_pval_threshold = 0.01
)

# RISKY: Uncorrected (exploratory only)
signatures_uncorrected <- discover_top_signatures(
  enrichment_data,
  fdr_correction = FALSE,
  fisher_pval_threshold = 0.05
)
```

### 2. Set Appropriate Thresholds

```{r best-practice-thresholds}
# Conservative (high confidence)
signatures_conservative <- discover_top_signatures(
  enrichment_data,
  fisher_pval_threshold = 0.001,    # Very stringent
  min_overlap_genes = 5,            # At least 5 genes
  min_cluster_presence = 5          # Must appear in 5+ clusters
)

# Moderate (recommended)
signatures_moderate <- discover_top_signatures(
  enrichment_data,
  fisher_pval_threshold = 0.01,
  min_overlap_genes = 3,
  min_cluster_presence = 3
)

# Liberal (exploratory)
signatures_liberal <- discover_top_signatures(
  enrichment_data,
  fisher_pval_threshold = 0.05,
  min_overlap_genes = 2,
  min_cluster_presence = 1
)
```

### 3. Validate Key Findings

```{r best-practice-validation}
# For top signatures, validate with:

# 1. Literature search
validate_with_literature <- function(pathway) {
  # Search PubMed for: "pathway" AND "Parkinson's disease"
  # Check if previously reported
}

# 2. Cross-cluster consistency
check_cross_cluster <- function(sig) {
  # Is signature present in multiple clusters?
  # Higher reproducibility = higher confidence
}

# 3. External datasets
validate_external <- function(sig) {
  # Check in other PD datasets
  # GSE series, Human Brain Atlas, etc.
}

# 4. Experimental validation
# - qPCR for key genes
# - Western blot for proteins
# - Functional assays
```

### 4. Document Analysis Parameters

```{r best-practice-documentation}
# Save complete analysis metadata
signature_metadata <- list(
  analysis_date = Sys.Date(),
  package_version = packageVersion("iSCORE.PDecipher"),
  enrichment_data_file = "all_enrichment_padj005_complete_with_direction.rds",
  parameters = list(
    top_n = 50,
    fisher_pval_threshold = 0.01,
    min_overlap_genes = 3,
    min_cluster_presence = 3,
    fdr_correction = TRUE,
    fdr_method = "BH"
  ),
  results_summary = list(
    total_signatures = nrow(signatures),
    high_confidence = sum(signatures$signature_score > 0.7),
    gene_pairs_analyzed = n_distinct(signatures$mast_gene, signatures$mixscale_gene)
  )
)

saveRDS(signature_metadata, "signature_analysis_metadata.rds")
```

## Troubleshooting

### Issue: "No signatures found"

**Causes:**
1. Thresholds too stringent
2. Poor gene overlap
3. Opposite effect directions (antagonistic)

**Solutions:**
```{r troubleshoot-no-sigs}
# 1. Relax thresholds
signatures <- discover_top_signatures(
  enrichment_data,
  fisher_pval_threshold = 0.05,  # Less stringent
  min_overlap_genes = 2           # Lower minimum
)

# 2. Check individual enrichments
# Are both methods finding enriched terms?

# 3. Examine effect directions
# Use direction_consistency analysis
```

### Issue: "Too many low-confidence signatures"

**Causes:**
1. Not using FDR correction
2. Large multiple testing burden

**Solutions:**
```{r troubleshoot-too-many}
# 1. Apply FDR correction
signatures <- discover_top_signatures(
  enrichment_data,
  fdr_correction = TRUE,
  fdr_method = "BH"
)

# 2. Increase stringency
signatures <- signatures %>%
  filter(signature_score > 0.7, fisher_pvalue < 0.001)

# 3. Focus on pan-cluster signatures
signatures <- signatures %>%
  group_by(pathway_description) %>%
  filter(n() >= 3) %>%
  ungroup()
```

## Conclusions

**Key Takeaways:**

1. ✅ Signature discovery identifies convergent pathways validated by two methods
2. ✅ High-confidence signatures (score > 0.7) are robust therapeutic targets
3. ✅ Pan-cluster signatures represent core disease mechanisms
4. ✅ PD-relevant pathways cluster into key categories (mitochondrial, autophagy, etc.)
5. ✅ Cross-method validation reduces false positives dramatically
6. ✅ Biological interpretation requires domain knowledge + statistical rigor

**Signature Discovery Strengths:**
- Robust to technical artifacts
- Cross-validated by independent methods
- Biologically interpretable
- Directly applicable to therapeutic development

**Next Steps:**
- Validate top signatures experimentally
- Explore cluster-specific mechanisms
- Integrate with other PD datasets
- Consider therapeutic interventions

## Additional Resources

- **Getting Started Vignette**: Basic iSCORE-PDecipher usage
- **Perturb-seq Workflow**: Detailed Mixscale integration
- **Advanced Customization**: Batch processing and automation
- **GitHub Issues**: Report bugs or request features

## Session Info

```{r session-info}
sessionInfo()
```

---

**Package**: iSCORE.PDecipher
**Feature**: Signature Discovery
**Algorithm**: Fisher's Exact Test + Effect Correlation + Composite Scoring
**Applications**: PD research, therapeutic target identification, pathway analysis
