---
title: "Parkinson's Disease Pathway Signatures: Cross-Method Analysis"
subtitle: "Executive Summary for Thesis Committee"
author: "iSCORE-PDecipher Analysis"
date: "July 19, 2025"
output: 
  html_document:
    toc: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(dplyr)
library(knitr)

# Load results
results_dir <- "../results/pd_signatures"
mast_top <- read.csv(file.path(results_dir, "mast_top_fast.csv"), stringsAsFactors = FALSE)
mixscale_top <- read.csv(file.path(results_dir, "mixscale_top_fast.csv"), stringsAsFactors = FALSE)
convergent_top <- read.csv(file.path(results_dir, "convergent_top_fast.csv"), stringsAsFactors = FALSE)
```

# Executive Summary

We identified **45,671 PD-relevant pathways** from 663,280 total enrichment results across iSCORE-PD mutations and CRISPRi knockdowns. Our analysis reveals three compelling signature categories with distinct biological insights into Parkinson's disease pathogenesis.

## Key Findings

### 1. **MAST-only Signature: Synaptic Vesicle Dysfunction**
- **Top pathway**: Vesicle-mediated transport in synapse (p < 10^-6)
- **Found in 8 genes**: LRRK2, PARK7, PRKN, and others
- **Biological insight**: Genetic mutations converge on synaptic vesicle trafficking defects

### 2. **CRISPRi-only Signature: Mitochondrial Compensation**  
- **Top pathway**: Mitochondrial small ribosomal subunit
- **Found in 9 genes**: ATP13A2, DNAJC6, FBXO7, and others
- **Biological insight**: Acute knockdowns trigger compensatory mitochondrial protein synthesis

### 3. **Convergent Signature: Universal Synaptic Dysfunction**
- **Top pathway**: Synapse (STRING database)
- **13 MAST genes, 10 CRISPRi genes**
- **Special finding**: "Dopamine perturbation network" shows extreme significance (p < 10^-17)
- **Biological insight**: Core synaptic dysfunction is a universal feature across all perturbation methods

# Detailed Analysis

## Method Comparison

```{r method_stats}
method_stats <- data.frame(
  Method = c("MAST (Mutations)", "CRISPRi (Knockdowns)", "Convergent"),
  `Total Pathways` = c(
    length(unique(mast_top$Description)),
    length(unique(mixscale_top$Description)),
    nrow(convergent_top)
  ),
  `Avg Genes/Pathway` = c(
    round(mean(mast_top$n_genes), 2),
    round(mean(mixscale_top$n_genes), 2),
    round(mean(convergent_top$n_genes_mast + convergent_top$n_genes_mixscale)/2, 2)
  ),
  `Top Significance` = c(
    round(max(mast_top$mean_neg_log_p), 2),
    round(max(mixscale_top$mean_neg_log_p), 2),
    round(max(convergent_top$mean_neg_log_p), 2)
  )
)

kable(method_stats, caption = "Summary statistics by analysis method")
```

## Top 5 Pathways by Category

### MAST-only Pathways
```{r mast_table}
mast_display <- mast_top %>%
  head(5) %>%
  select(Pathway = Description, 
         `Genes (n)` = n_genes,
         `Example Genes` = genes,
         `-log10(p)` = mean_neg_log_p) %>%
  mutate(
    Pathway = substr(Pathway, 1, 60),
    `-log10(p)` = round(`-log10(p)`, 2)
  )

kable(mast_display, caption = "Top MAST-only pathways")
```

### CRISPRi-only Pathways
```{r mixscale_table}
mixscale_display <- mixscale_top %>%
  head(5) %>%
  select(Pathway = Description,
         `Genes (n)` = n_genes,
         `Example Genes` = genes,
         `-log10(p)` = mean_neg_log_p) %>%
  mutate(
    Pathway = substr(Pathway, 1, 60),
    `-log10(p)` = round(`-log10(p)`, 2)
  )

kable(mixscale_display, caption = "Top CRISPRi-only pathways")
```

### Convergent Pathways
```{r convergent_table}
convergent_display <- convergent_top %>%
  head(5) %>%
  select(Pathway = Description,
         `MAST Genes` = n_genes_mast,
         `CRISPRi Genes` = n_genes_mixscale,
         `-log10(p)` = mean_neg_log_p) %>%
  mutate(
    Pathway = substr(Pathway, 1, 60),
    `-log10(p)` = round(`-log10(p)`, 2)
  )

kable(convergent_display, caption = "Top convergent pathways (found in both methods)")
```

# Biological Interpretation

## 1. Synaptic Dysfunction as Central Theme
- **MAST mutations** specifically affect vesicle trafficking
- **Convergent analysis** confirms broad synaptic involvement
- **Dopamine-specific** network highly enriched (p < 10^-17)

## 2. Method-Specific Insights
- **Mutations (MAST)**: Chronic adaptations → vesicle/transport defects
- **Knockdowns (CRISPRi)**: Acute stress → mitochondrial compensation

## 3. Therapeutic Implications
- **Synaptic vesicle pathway** genes are druggable targets
- **Mitochondrial support** may help in acute phases
- **Dopamine network** components for biomarker development

# Recommendations for Manuscript

## Main Figure Components
1. **Three-panel bar plot** showing top pathways by category
2. **Heatmap** of convergent pathways across all genes
3. **Network diagram** connecting PD genes through shared pathways

## Key Messages
1. **Validation**: Convergent signatures validate our dual-method approach
2. **Novel insights**: Method-specific signatures reveal different aspects of PD biology
3. **Clinical relevance**: Identified pathways suggest therapeutic strategies

## Statistical Highlights
- 934 convergent pathways (p < 0.05 in both methods)
- Dopamine perturbation network: p < 10^-17
- 13 PD genes converge on synaptic pathways

# Next Steps

1. **Deep dive** into dopamine perturbation network genes
2. **Validate** key findings in additional datasets
3. **Explore** therapeutic compounds targeting identified pathways
4. **Develop** pathway-based biomarker panel

---

*Generated by iSCORE-PDecipher v0.2.9 | Data: iSCORE-PD + CRISPRi (no CRISPRa)*